---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# erplots

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

erplots provides a fluent mini-language for building exposure-response
plots: model curves/ribbons, quantile-binned response-rate summaries, data
strips, and grouped distribution panels. It is model-agnostic: erplots
never fits a model itself. Instead, you fit a model with whatever package
suits your workflow (e.g. [erlr](https://github.com/djnavarro/erlr) for
logistic regression), and pass the fitted object to `er_plot_show_model()`.
Any model that implements `er_predict()` can be visualised; implementing
`er_simulate()` and `er_summary()` additionally enables uncertainty
spaghetti plots/VPCs and summary annotations (e.g. p-values). See
`?er_model_interface`.

## Installation

You can install the development version of erplots like so:

``` r
pak::pak("djnavarro/erplots")
```

## Example

```{r}
#| fig-height: 8
#| label: er-plot

library(erplots)
library(erlr)

mod <- lr_model(ae1 ~ aucss, lr_data)

lr_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_groups(aucss) |> 
  plot()

mod2 <- lr_model(ae2 ~ aucss + sex, lr_data)
mod2_marginal <- lr_model(ae2 ~ aucss, lr_data)

plt <- lr_data |> 
   er_plot(aucss, ae2, stratify_by = sex) |> 
   # keep_strata = FALSE needs a model without the stratification
   # variable as a term, so we pass `mod2_marginal` here
   er_plot_show_model(mod2_marginal, keep_strata = FALSE) |> 
   er_plot_show_quantiles(bins = 3) |> 
   er_plot_show_datastrip() |> 
   er_plot_show_groups(group_by = c(aucss, treatment), keep_strata = FALSE)

print(plt)
plot(plt)
```

## VPC plots

```{r}
#| label: er-vpc
sim <- lr_vpc_sim(mod2, seed = 1234)
sim

er_vpc_plot(lr_data, sim, aucss, ae2, group_by = aucss)
er_vpc_plot(lr_data, sim, aucss, ae2, group_by = sex)
```
