---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# erplots

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/djnavarro/erplots/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/djnavarro/erplots/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/djnavarro/erplots/graph/badge.svg)](https://app.codecov.io/gh/djnavarro/erplots)
<!-- badges: end -->

The erplots package provides a mini-language for building exposure-response
plots: model curves/ribbons, quantile-binned response-rate summaries, data
strips, and grouped distribution panels. It is model-agnostic: erplots
never fits a model itself. Instead, you fit a model with whatever package
suits your workflow (e.g. [erglm](https://github.com/djnavarro/erglm) for
logistic regression), and pass the fitted object to `er_plot_add_model()`.

Any model that implements `er_predict()` can be visualised; implementing
`er_simulate()` and `er_summary()` additionally enables uncertainty
spaghetti plots/VPCs and summary annotations (e.g. p-values). See the
[Implementing the model interface](https://erplots.djnavarro.net/articles/model-interface.html)
article for a worked guide to implementing all three, or
`?er_model_interface` for the bare contract.

## Installation

You can install the development version of erplots like so:

``` r
pak::pak("djnavarro/erplots")
```

## Example

```{r}
#| fig-height: 8
#| label: er-plot

library(erplots)
library(erglm)

mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_groups(aucss) |> 
  plot()

mod2 <- erglm_model(ae2 ~ aucss + sex, erglm_data, family = binomial())

plt <- erglm_data |> 
   er_plot(aucss, ae2, stratify_by = sex) |> 
   er_plot_add_model(mod2) |> 
   er_plot_add_quantiles(bins = 3) |> 
   er_plot_add_data() |> 
   er_plot_add_groups(group_by = c(aucss, treatment), keep_strata = FALSE)

print(plt)
plot(plt)
```
