---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# erplots

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/djnavarro/erplots/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/djnavarro/erplots/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/djnavarro/erplots/graph/badge.svg)](https://app.codecov.io/gh/djnavarro/erplots)
[![CRAN status](https://www.r-pkg.org/badges/version/erplots)](https://CRAN.R-project.org/package=erplots)
<!-- badges: end -->

The erplots package provides a mini-language for building exposure-response
plots: model curves/ribbons, quantile-binned response-rate summaries, data
strips, and grouped distribution panels. It supports time-to-event plots and
visual predictive check plots as well as the core exposure-response plots,
and it is model agnostic: it can display model predictions and summaries for
a particular model type as long an appropriate model interface (consisting of 
a few key S3 methods) is available. Packages that support the interface
include [erglm](https://erglm.djnavarro.net) (e.g., logistic regression,
linear regression, Poisson regression, etc), [ertte](https://ertte.djnavarro.net)
(parametric survival models, Cox proportional hazard models), and 
[emaxnls](https://emaxnls.djnavarro.net) (hyperbolic and sigmoidal Emax 
regression models for binary and continuous outcomes).

## Installation

You can install the latest CRAN release like so:

``` r
install.packages("erplots")
```

Alternatively, you can install the development version of erplots with this:

``` r
pak::pak("djnavarro/erplots")
```

## Example

```{r}
#| fig-height: 8
#| label: er-plot

library(erplots)
library(erglm)

mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_groups(aucss) |> 
  plot()

mod2 <- erglm_model(ae2 ~ aucss + sex, erglm_data, family = binomial())

plt <- erglm_data |> 
   er_plot(aucss, ae2, stratify_by = sex) |> 
   er_plot_add_model(mod2) |> 
   er_plot_add_quantiles(bins = 3) |> 
   er_plot_add_data() |> 
   er_plot_add_groups(group_by = c(aucss, treatment), keep_strata = FALSE)

print(plt)
plot(plt)
```
