---
title: "The plotting grammar"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article is about the *grammar* erplots plots are built from, not
about any one plot's mechanics -- for worked examples of each layer, see
the [Plotting](plot.html) article. If a design choice described here
ever changes, this article and `PLAN.md`'s "Mini-language architecture
review" section should be updated together; see the note at the end.

```{r setup}
library(erplots)
library(erglm)
```

## Building a plot is composing layers onto an object

`er_plot()` creates an empty object of class `er_plot`, storing the data
and the plot's exposure/response/stratification variables. You then pipe
it through one or more *layer* functions, each of which adds one visual
component, and finish with `plot()`/`print()` (or [er_plot_build()]
directly):

```{r pipeline}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_model(mod) |>
  er_plot_show_quantiles() |>
  er_plot_show_groups(aucss) |>
  plot()
```

```
observed data
      |
      v
  er_plot()                       -- creates the (empty) object
      |
      v
  layer functions (piped, any order, any subset):
    er_plot_show_model()
    er_plot_show_quantiles()
    er_plot_show_datastrip()
    er_plot_show_groups()
      |
      v
  er_plot_build()                 -- called for you by plot()/print()
      |
      v
  polish: margins, labels, legends, theme
      |
      v
  compose: patchwork
      |
      v
  rendered plot
```

There are currently four layers, each documented on its own help topic:

| Layer | Function | Shows | Depends on `response_type`? |
|---|---|---|---|
| Model | [er_plot_show_model()] | Fitted curve/ribbon (or spaghetti) plus an optional summary annotation | No |
| Quantile | [er_plot_show_quantiles()] | Exposure-quantile-binned response summary (rate/mean + CI) | Yes |
| Data strip | [er_plot_show_datastrip()] | Raw observations jittered along exposure | Yes (binary only, for now) |
| Group | [er_plot_show_groups()] | Exposure distribution, boxplot/violin, split by a grouping variable | No |

## Layers are either singleton or additive

Calling a layer function twice on the same object doesn't always do the
same thing. The model, quantile, and data-strip layers are
**singleton**: a second call replaces the first call's result rather
than combining the two.

```{r singleton}
plt <- erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_quantiles(bins = 4) |>
  er_plot_show_quantiles(bins = 8) # overwrites the bins = 4 call

plt$part$quantile$config$n_quantiles # 8, not 4
```

The group layer is the one exception: it's **additive**. Each call adds
another panel alongside any already added, rather than replacing them:

```{r additive}
plt <- erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_groups(aucss) |>
  er_plot_show_groups(treatment) # adds a second panel, doesn't replace the first

names(plt$part$group$config) # both grouping variables are still there
```

This is a deliberate design choice, not an oversight: there is only one
"the model" and one "the quantile summary" to show per plot, but many
legitimate ways to slice the exposure distribution by different grouping
variables. The one flagged exception to watch for: because the model
layer is singleton, overlaying two model curves for comparison (e.g. a
candidate model against a reference model) isn't currently possible.
That's tracked as a plausible future addition, not planned work -- see
`PLAN.md`.

## Stratification composes with layers, usually via color

`stratify_by`, set once in `er_plot()`, declares a single discrete
variable used to split layers by color/fill, with one shared,
deduplicated legend across the whole composed plot:

```{r stratification}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_show_model(mod_strat) |>
  er_plot_show_quantiles() |>
  er_plot_show_datastrip() |>
  plot()
```

Each layer's own `keep_strata` argument controls whether *that* layer
uses the stratification (default `TRUE` whenever `stratify_by` was set).
The general rule, in the order a layer actually applies it: **a layer's
own encoding takes precedence; stratification adapts to whatever channel
is left**, defaulting to color/fill. Today, every layer's own encoding
leaves color/fill free for strata, so this rule is invisible in practice
-- but it stops being invisible once a layer needs color for something
else. The data strip layer is the first anticipated case: a
continuous-response variant would need color for the response value
itself, so stratification for that one layer would have to fall back to
per-stratum facets instead. That variant doesn't exist yet (see
`PLAN.md`'s "Continuous-response data strip" section); today's data
strip only supports a binary response, and uses color/fill for strata
like every other layer.

## Response type changes what a layer summarises, not whether it appears

`response_type`, also set once in `er_plot()` (`"auto"`, `"binary"`,
`"continuous"`, or `"count"`), governs the response's scale and which
summary/CI method a response-type-aware layer uses. The model and group
layers don't look at the response's type at all -- they only consume
[er_predict()]/[er_simulate()] output or the exposure variable,
respectively. The quantile layer dispatches on it directly:

| `response_type` | Bin summary | CI method |
|---|---|---|
| `"binary"` | Response rate | Clopper-Pearson ([clopper_pearson()]) |
| `"continuous"` | Mean | t-interval ([t_interval()]) |
| `"count"` | Mean | Exact Poisson interval ([poisson_interval()]) |

`"auto"` classifies a response as `"binary"` if it's logical or confined
to `{0, 1}`, and `"continuous"` otherwise -- so a genuine count response
auto-detects as `"continuous"` and is summarised as an
approximately-continuous quantity unless `response_type = "count"` is
declared explicitly. See [er_plot_show_quantiles()] for the full
rationale and the [Plotting](plot.html) article's "Continuous responses"
section for worked examples, including a case where the choice between
the t-interval and exact Poisson interval visibly matters.

The data strip layer is the odd one out: rather than adapting its
summary statistic, it currently only supports `"binary"` responses at
all, and errors for `"continuous"`/`"count"` rather than silently
mis-plotting -- see [er_plot_show_datastrip()].

## Keeping this article in sync

This article describes the grammar as designed and implemented *today*.
Whenever a future change lands that alters any of the above --
renaming a layer, changing which layers are singleton vs. additive,
changing how stratification composes with a layer's own encoding, or
adding/removing a `response_type` dispatch -- update this article in the
same change, and update the cross-references in `PLAN.md`'s
"Mini-language architecture review" section (which points here) to match.
Treat a design change that isn't reflected here as incomplete.
