---
title: "The plotting grammar"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article is about the *grammar* erplots plots are built from, not
about any one plot's mechanics -- for worked examples of each layer, see
the [Plotting](plot.html) article. If a design choice described here
ever changes, this article and `PLAN.md`'s "Mini-language architecture
review" section should be updated together; see the note at the end.

```{r setup}
library(erplots)
library(erglm)
```

## Building a plot is composing layers onto an object

`er_plot()` creates an empty object of class `er_plot`, storing the data
and the plot's exposure/response/stratification variables. You then pipe
it through one or more *layer* functions, each of which adds one visual
component, and finish with `plot()`/`print()` (or [er_plot_build()]
directly):

```{r pipeline}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_model(mod) |>
  er_plot_show_quantiles() |>
  er_plot_show_groups(aucss) |>
  plot()
```

```
observed data
      |
      v
  er_plot()                       -- creates the (empty) object
      |
      v
  layer functions (piped, any order, any subset):
    er_plot_show_model()
    er_plot_show_quantiles()
    er_plot_show_data()
    er_plot_show_groups()
      |
      v
  er_plot_build()                 -- called for you by plot()/print()
      |
      v
  polish: margins, labels, legends, theme
      |
      v
  compose: patchwork
      |
      v
  rendered plot
```

There are currently four layers, each documented on its own help topic:

| Layer | Function | Shows | Depends on `response_type`? |
|---|---|---|---|
| Model | [er_plot_show_model()] | Fitted curve/ribbon (or spaghetti) plus an optional summary annotation | No |
| Quantile | [er_plot_show_quantiles()] | Exposure-quantile-binned response summary (rate/mean + CI) | Yes |
| Data | [er_plot_show_data()] | Raw observations, by default overlaid on the model panel at their true (exposure, response) coordinates ([build_data_overlay()]); or, for a binary response, [build_data_boxjitter()]'s older panel-based boxplot + jitter design | Yes |
| Group | [er_plot_show_groups()] | Exposure distribution, boxplot/violin, split by a grouping variable | No |

## Layers are either singleton or additive

Calling a layer function twice on the same object doesn't always do the
same thing. The model, quantile, and data layers are
**singleton**: a second call replaces the first call's result rather
than combining the two.

```{r singleton}
plt <- erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_quantiles(bins = 4) |>
  er_plot_show_quantiles(bins = 8) # overwrites the bins = 4 call

plt$part$quantile$config$n_quantiles # 8, not 4
```

The group layer is the one exception: it's **additive**. Each call adds
another panel alongside any already added, rather than replacing them:

```{r additive}
plt <- erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_groups(aucss) |>
  er_plot_show_groups(treatment) # adds a second panel, doesn't replace the first

names(plt$part$group$config) # both grouping variables are still there
```

This is a deliberate design choice, not an oversight: there is only one
"the model" and one "the quantile summary" to show per plot, but many
legitimate ways to slice the exposure distribution by different grouping
variables. The one flagged exception to watch for: because the model
layer is singleton, overlaying two model curves for comparison (e.g. a
candidate model against a reference model) isn't currently possible.
That's tracked as a plausible future addition, not planned work -- see
`PLAN.md`.

## Stratification composes with layers, usually via color

`stratify_by`, set once in `er_plot()`, declares a single discrete
variable used to split layers by color/fill, with one shared,
deduplicated legend across the whole composed plot:

```{r stratification}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_show_model(mod_strat) |>
  er_plot_show_quantiles() |>
  er_plot_show_data() |>
  plot()
```

Each layer's own `keep_strata` argument controls whether *that* layer
uses the stratification (default `TRUE` whenever `stratify_by` was set).
The general rule, in the order a layer actually applies it: **a layer's
own encoding takes precedence; stratification adapts to whatever channel
is left**, defaulting to color/fill.

For most layers, color/fill is always free for strata, so this rule is
invisible in practice. The data layer is the one exception, and its
behaviour now depends on which builder is in play, and which
*structural* family (declared via [er_layout()]) that builder belongs
to:

- `build_data_overlay()` (the default, `"overlay"`-layout): color, when
  mapped at all, always means strata -- the response is already shown
  via y-position, so color/fill is free for stratification like every
  other layer, and the overlay shares the base plot's own strata legend
  with the model/quantile layers.
- `build_data_boxjitter()` (the older, panel-based design,
  `"panel"`-layout, binary-response only): behaves the same way as
  overlay -- color/fill means strata, shared legend. There is no
  built-in `"panel"`-layout builder for a continuous/count response
  today; if one is written, its color aesthetic would typically already
  be spoken for by the response value itself (as the removed
  `build_data_color()` builder's was), in which case stratification
  should fall back to one panel per stratum level instead of a shared
  legend -- the concrete instance of "a layer's own encoding takes
  precedence" that motivated the general rule. See `PLAN.md`'s
  "Continuous-response data strip" section for that design history, and
  [er_plot_show_data()] for the full breakdown.

A `config$color_role` tag (`"strata"` or `"response"`, set by
`.part_data()`) records which meaning applies for a given data-layer
build, so the composition machinery (`.polish_labels()`/
`.polish_legends()`) knows whether to treat a builder's legend as the
shared strata legend or a standalone response colorbar.

## Response type changes what a layer summarises, not whether it appears

`response_type`, also set once in `er_plot()` (`"auto"`, `"binary"`,
`"continuous"`, or `"count"`), governs the response's scale and which
summary/CI method a response-type-aware layer uses. The model and group
layers don't look at the response's type at all -- they only consume
[er_predict()]/[er_simulate()] output or the exposure variable,
respectively. The quantile layer dispatches on it directly:

| `response_type` | Bin summary | CI method |
|---|---|---|
| `"binary"` | Response rate | Clopper-Pearson ([clopper_pearson()]) |
| `"continuous"` | Mean | t-interval ([t_interval()]) |
| `"count"` | Mean | Exact Poisson interval ([poisson_interval()]) |

`"auto"` classifies a response as `"binary"` if it's logical or confined
to `{0, 1}`, and `"continuous"` otherwise -- so a genuine count response
auto-detects as `"continuous"` and is summarised as an
approximately-continuous quantity unless `response_type = "count"` is
declared explicitly. See [er_plot_show_quantiles()] for the full
rationale and the [Plotting](plot.html) article's "Continuous responses"
section for worked examples, including a case where the choice between
the t-interval and exact Poisson interval visibly matters.

The data layer doesn't compute a summary statistic at all -- it just
plots raw observations -- so `response_type` instead changes *how* it's
drawn: `build_data_overlay()` (the default) needs no dispatch (a plain
scatter, or a small vertical jitter for a binary response's exactly-0/1
y-values); `build_data_boxjitter()` is binary-response only, and uses
`response_type` only insofar as `er_plot_show_data()` guards against
using it on a continuous/count response at all (there's no upper/lower
partition to split on) -- see [er_plot_show_data()].

## Extending erplots: writing your own builder

Every layer function delegates the actual drawing to a `build_*()`
function sharing a common signature -- `function(data, config, stratify,
exposure, response, strata, style)`. That signature is a documented,
public part of the API (see [er_partial()]), and each layer function's
`builder` argument (`er_plot_show_model()` additionally has
`summary_builder`) defaults to one built-in `build_*()` function (e.g.
[er_plot_show_quantiles()]'s defaults to `build_quantile_errorbar()`)
and can be set to any other function matching the same signature -- no
need to fork the package or reach into `object$part` internals:

```{r builder}
build_quantile_crossbar <- function(data, config, stratify, exposure, response, strata, style) {
  ggplot2::geom_crossbar(
    data = config$summary,
    mapping = ggplot2::aes(x = x_mid, y = y_mid, ymin = ci_lower, ymax = ci_upper),
    inherit.aes = FALSE
  )
}

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_show_model(mod) |>
  er_plot_show_quantiles(builder = build_quantile_crossbar) |>
  plot()
```

A custom builder receives the same pre-computed `config` a built-in
builder would (e.g. `config$summary` for the quantile layer,
`config$predictions` for the model layer), so it only needs to turn that
`config` into ggplot2 layers, not recompute anything.

For the data layer, a custom `builder` must also declare which
*structural* family it belongs to, via [er_layout()]: `er_layout(fn,
"overlay")` slots it into a single call merged onto the main panel,
while `er_layout(fn, "panel")` slots it into one-or-more panels stacked
below the base plot. [er_plot_show_data()] reads this tag off `builder`
itself to decide how to assemble the layer, rather than taking a
separate argument for it -- so a builder like `build_data_overlay()` can
never accidentally end up routed into upper/lower panels, or vice versa.
The other three layers have only one structural call site, so no such
tagging is needed there.

`build_quantile_pointrange()` (a single `geom_pointrange()` in place of
`build_quantile_errorbar()`'s separate point + error bar) started life
as exactly this kind of custom builder, and was promoted to a built-in
option once it proved to need no new `config` fields -- a reasonable bar
to check your own custom builders against, if you're deciding whether to
propose one upstream. See the `@examples` on [er_plot_show_model()],
[er_plot_show_quantiles()], and [er_plot_show_data()] for further worked
custom builders (a dashed model curve and a data-overlay scatter), and
[er_partial()]'s "Writing your own builder" section for the full
contract.

## Keeping this article in sync

This article describes the grammar as designed and implemented *today*.
Whenever a future change lands that alters any of the above --
renaming a layer, changing which layers are singleton vs. additive,
changing how stratification composes with a layer's own encoding, or
adding/removing a `response_type` dispatch -- update this article in the
same change, and update the cross-references in `PLAN.md`'s
"Mini-language architecture review" section (which points here) to match.
Treat a design change that isn't reflected here as incomplete.
