---
title: "Understanding the erplots grammar"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of this article is to describe the **grammar** that erplots uses to generate exposure-response plots. It's not intended to describe all the style options available to users, or cover the mechanics of how plots are constructed. If you want to see those in more detail, there are articles covering [binary](plot-binary.html), [continuous](plot-continuous.html), and
[count](plot-count.html) response data. You can also read the article on [extending erplots](extending.html) if you want to learn how you can write your own functions that change how the plots are displayed. 

<!-- agent-note: whenever a future change lands that alters any of the
above -- renaming a layer, changing which layers are singleton vs.
additive, changing how stratification composes with a layer's own
encoding, or adding/removing a `response_type` dispatch -- update this
article in the same change. Treat a design change that isn't reflected
here as incomplete. -->


```{r setup}
library(erplots)
library(erglm)
```

## Plots are constructed in layers

The design of an exposure-response plot follows a very similar logic to how plots are built in the ggplot2 package. The first function you call is always `er_plot()`, which creates an empty object of class `er_plot`. When you do this, it stores the data set internally and keeps track of which columns correspond to the exposure, response, and stratification variables. You then pipe it through through one or more *layer* functions, each of which adds one visual layer, and then finish by calling `plot()` to generate the plot. A typical plot generation pipeline might look like this:

```{r pipeline}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles() |>
  er_plot_add_groups(aucss) |>
  plot()
```

There are a few features to notice here. 

- First, the model object `mod` is not created by the erplots package itself. In this case, we fit the model using `erglm::erglm_model()`, and because the erglm package implements the erplots model interface functions, the erplots package knows how to extract everything it needs from the model.
- Second, notice that at the end of the pipeline we call `plot()` explicitly.One difference between ggplot2 and erplots is that erplots makes a strong distinction between calling `print()` on the plot object (outputs a written description of the plot) and calling `plot()` on the plot object (renders the plot). 
- Third, although erplots uses the "layers" language to talk about different parts of the plot, an erplot layer operates at a different level of abstraction to a ggplot2 geom. A single erplot layer might add several ggplot2 geoms rather than a single one; it might even create a separate ggplot2 object that gets merged with other ggplot2 objects using the patchwork package. 

The diagram below illustrates the structure of the erplots plotting pipeline:


```
observed data
      |
      v
  er_plot()                       -- creates the (empty) object
      |
      v
  layer functions (piped, any order, any subset):
    er_plot_add_model()
    er_plot_add_summary()
    er_plot_add_quantiles()
    er_plot_add_data()
    er_plot_add_groups()
      |
      v
  er_plot_build()                 -- called for you by plot()
      |
      v
  polish: margins, labels, legends, theme
      |
      v
  compose: patchwork
      |
      v
  rendered plot
```

From the user perspective, everything ends with the `er_plot_build()` function, which you don't generally call directly: passing your plot to the `plot()` function will do that part for you. The additional "polish" and "compose" steps are handled behind the scenes, but shown in the diagram to give you a sense of what happens. 

There are currently five different kinds of layer supported by erplots, each documented on its own help topic:

| Layer | Function | Shows |
|---|---|------|
| Model | `er_plot_add_model()` | Fitted curve/ribbon (or spaghetti plot) that shows what the model predicts |
| Summary | `er_plot_add_summary()` | A corner-placed annotation describing some aspect to the model (e.g., p-value) or the data set (e.g., sample statistics) |
| Quantile | `er_plot_add_quantiles()` | Exposure-quantile-binned response summary, usually a mean and confidence interval |
| Data | `er_plot_add_data()` | Raw observations, by default overlaid on the model panel at their true (exposure, response) coordinates, but sometimes added to separate "data strips" that appear above and below the model panel |
| Group | `er_plot_add_groups()` | Exposure distribution, boxplot/violin, split by a grouping variable |

## Layers are singleton or additive

Calling a layer function twice on the same object doesn't always do the
same thing. The model, summary, quantile, and data layers are
**singleton**: a second call replaces the first call's result rather
than combining the two.

```{r singleton}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_quantiles(bins = 8) |>
  er_plot_add_quantiles(bins = 4) |> # overwrites the bins = 8 call
  plot() # rendered plot has 4 quantile bins (plus placebo group)
```

The group layer is the one exception: it's **additive**. Each call adds
another panel alongside any already added, rather than replacing them:

```{r additive}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_groups(aucss) |> # adds first panel
  er_plot_add_groups(treatment) |> # adds second panel
  plot() # rendered plot contains both
```

This is a deliberate design choice, not an oversight: there is only one
"the model" and one "the quantiles" to show per plot, but many
legitimate ways to slice the exposure distribution by different grouping
variables. 

Note that there is one important side-effect to this: because the model
layer is singleton, overlaying two model curves for comparison (e.g. a
candidate model against a reference model) isn't currently possible.
That's a plausible future addition, but not something currently planned.

## Stratification composes with layers

`stratify_by`, set once in `er_plot()`, declares a single discrete
variable used to split layers by color/fill, with one shared,
deduplicated legend across the whole composed plot:

```{r stratification}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  plot()
```

Each layer's own `keep_strata` argument controls whether *that* layer
uses the stratification (default `TRUE` whenever `stratify_by` was set).
The general rule, in the order a layer actually applies it: **a layer's
own encoding takes precedence; stratification adapts to whatever channel
is left**, defaulting to color/fill.

For most layers, color/fill is always free for strata, so this rule is
invisible in practice. The data layer is the one exception, and its
behaviour now depends on which builder is in play, and which
*structural* family (declared via `er_style_tag()`) that builder belongs
to:

- `er_style_data_overlay()` (the default, `"overlay"`-layout): color, when
  mapped at all, always means strata -- the response is already shown
  via y-position, so color/fill is free for stratification like every
  other layer, and the overlay shares the base plot's own strata legend
  with the model/quantile layers.
- `er_style_data_boxjitter()` (the older, panel-based design,
  `"panel"`-layout, binary-response only): behaves the same way as
  overlay -- color/fill means strata, shared legend. There is no
  built-in `"panel"`-layout builder for a continuous/count response
  today; if one is written, its color aesthetic would typically already
  be spoken for by the response value itself, in which case
  stratification should fall back to one panel per stratum level
  instead of a shared legend -- the concrete instance of "a layer's own
  encoding takes precedence" that motivated the general rule. See
  [er_plot_add_data()] for the full breakdown.

A `config$color_role` tag (`"strata"` or `"response"`) records which
meaning applies for a given data-layer build, so erplots knows whether
to treat a builder's legend as the shared strata legend or a standalone
response colorbar.

## Response type changes what a layer summarises

`response_type`, also set once in `er_plot()` (`"auto"`, `"binary"`,
`"continuous"`, or `"count"`), governs the response's scale and which
summary/CI method a response-type-aware layer uses. The model and group
layers don't look at the response's type at all -- they only consume
`er_predict()`/`er_simulate()` output or the exposure variable,
respectively. The quantile layer dispatches on it directly:

| `response_type` | Bin summary | CI method |
|---|---|---|
| `"binary"` | Response rate | Clopper-Pearson (`ci_clopper_pearson()`) |
| `"continuous"` | Mean | t-interval (`ci_t()`) |
| `"count"` | Mean | Exact Poisson interval (`ci_poisson()`) |

`"auto"` classifies a response as `"binary"` if it's logical or confined
to `{0, 1}`, and `"continuous"` otherwise -- so a genuine count response
auto-detects as `"continuous"` and is summarised as an
approximately-continuous quantity unless `response_type = "count"` is
declared explicitly. See `er_plot_add_quantiles()` for the full
rationale and the [count responses](plot-count.html) article's "Quantile
layer" section for a worked example, including a case where the
choice between the t-interval and exact Poisson interval visibly
matters.

The data layer doesn't compute a summary statistic at all -- it just
plots raw observations -- so `response_type` instead changes *how* it's
drawn: `er_style_data_overlay()` (the default) needs no dispatch (a plain
scatter, or a small vertical jitter for a binary response's exactly-0/1
y-values); `er_style_data_boxjitter()` is binary-response only, and uses
`response_type` only insofar as `er_plot_add_data()` guards against
using it on a continuous/count response at all (there's no upper/lower
partition to split on) -- see `er_plot_add_data()`.

## Theming is a separate concern

Everything above is about the **grammar**: which layers exist, whether a
second call to a layer function replaces or adds to the first, and how
stratification/`response_type` change what a layer draws. None of it
touches *how the composed plot looks* -- labels, axis limits, the overall
ggplot2 theme, the discrete color/fill palette, formatters, and so on.
That's a deliberately separate, orthogonal knob, `er_plot_theme()`,
covered in its own article: [Theming erplots](theming.html).

## Extending erplots

Writing a custom builder in detail -- including what `config` actually
contains for each layer, a worked crossbar example, and
`er_style_tag()`, the single helper a builder can use to declare its
`layout`/`fill_role`/`y_role` metadata for the composition machinery --
is its own article:
[Extending erplots](extending.html).

