---
title: "Extending erplots: writing your own builder"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article is about writing a custom `er_builder_*()` function -- the
mechanism by which any of erplots' four layers can be drawn differently
from its built-in options, without forking the package. It assumes
you're already familiar with the plotting grammar described in
[the plot grammar article](design.html) (layers, singleton/additive
semantics, stratification); this article goes into more depth on the
one topic that article only introduces: the `builder` escape hatch
itself.

```{r setup}
library(erplots)
library(erglm)
```

## The builder signature

Every layer function (`er_plot_add_model()`, `er_plot_add_quantiles()`,
`er_plot_add_data()`, `er_plot_add_groups()`) delegates its actual
drawing to a `builder` argument (`er_plot_add_model()` additionally has
`summary_builder`), which defaults to one built-in `er_builder_*()`
function and can be set to any other function -- built-in or custom --
sharing this signature:

```r
function(data, config, stratify, exposure, response, strata, style)
```

| Argument | What it is |
|---|---|
| `data` | The original data frame passed to `er_plot()`, unmodified. |
| `config` | The pre-computed configuration for *this specific layer* -- see below. |
| `stratify` | `TRUE`/`FALSE`: whether this layer should honour `stratify_by`. |
| `exposure`, `response`, `strata` | Plot-variable metadata lists (`name`, `label`, `limits`, ...) describing the exposure, response, and stratification variables declared in `er_plot()`. |
| `style` | Shared styling helpers: `style$theme_base()`, `style$draw_key`, `style$format_percent()`, `style$format_number()`. |

The function returns a geom, or a list of geoms/other objects that can
be added to a ggplot2 plot -- nothing more. This signature is documented
as public API on `?er_partial`, alongside each layer's own
`er_builder_*()` family page (`?er_builder_model`, `?er_builder_quantile`,
etc.).

### `config` is the part that matters, and it's already computed for you

`config` is where a custom builder actually gets its data from, and it
is **not** the raw `data` frame -- it's whatever the corresponding
internal `.part_*()` function derived from `data`/`exposure`/`response`/
`strata` before any builder ran. A custom builder's whole job is to turn
that already-computed `config` into ggplot2 layers; it never needs to
re-bin, re-summarise, or re-fit anything itself. Concretely:

| Layer | `.part_*()` | Key `config` field | Contents |
|---|---|---|---|
| Model | `.part_model()` | `config$predictions` | One row per exposure grid point, with `fit_resp`, `ci_lower`, `ci_upper` (from `er_predict()`) |
| Quantile | `.part_quantile()` | `config$summary` | One row per exposure-quantile bin (× stratum), with `x_mid`, `y_mid`, `ci_lower`, `ci_upper`, plus label-placement columns |
| Data | `.part_data()`/`.part_overlay()` | (none extra) | The builder mostly works from `data` directly, since this layer draws raw observations rather than a summary |
| Group | `.part_group()` | `config[[group_var]]$data`, `config[[group_var]]$counts` | The subset of `data` for that grouping variable, joined to per-group sample-size labels |

## Worked example: a custom quantile builder

Suppose the built-in `er_builder_quantile_errorbar()` (point + error bar)
and `er_builder_quantile_pointrange()`/`er_builder_quantile_bar()`
alternatives all feel like the wrong idiom, and you'd rather draw the
per-bin summary as a `geom_crossbar()`. First, look at what
`config$summary` actually contains, by building the quantile part on its
own and inspecting it -- this is the step a custom builder's author
does once, by hand, before writing the builder:

```{r inspect-config}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

plt <- erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_quantiles(bins = 6)

plt$part$quantile$config$summary
```

`x_mid`/`y_mid` are the bin's mean exposure and response rate;
`ci_lower`/`ci_upper` are the Clopper-Pearson interval bounds a builder
would map to `ymin`/`ymax`. (The other columns --
`y_mid_lbl`/`y_lwr_lbl`/`y_upr_lbl`/`y_lbl` -- support the built-in
label geom that `er_builder_quantile_errorbar()` draws alongside its error
bar; a builder that skips labels, like the one below, can ignore them.)
With that in hand, the builder itself is a small function that maps
those columns onto `geom_crossbar()`'s aesthetics:

```{r builder}
er_builder_quantile_crossbar <- function(data, config, stratify, exposure, response, strata, style) {
  ggplot2::geom_crossbar(
    data = config$summary,
    mapping = ggplot2::aes(x = x_mid, y = y_mid, ymin = ci_lower, ymax = ci_upper),
    inherit.aes = FALSE
  )
}

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles(builder = er_builder_quantile_crossbar) |>
  plot()
```

A few things worth noting about this builder, all generalisable to any
layer:

- It ignores `data`, `stratify`, `exposure`, `response`, `strata`, and
  `style` entirely -- a builder only needs to use the arguments relevant
  to what it draws. (A stratified version would need to map
  `color = strata` and use `config$summary`'s `strata` column, plus
  `style$draw_key` for a legend key consistent with the other layers --
  see `er_builder_quantile_errorbar()`'s source for a worked stratified
  example.)
- `inherit.aes = FALSE` is there because this geom supplies its own
  `data`/`mapping`, distinct from whatever's already on the base plot
  (the model curve, in this example) -- omitting it would try to inherit
  the base plot's aesthetics and fail, since those don't include
  `ymin`/`ymax`.
- No new `config` fields were needed. `er_builder_quantile_pointrange()`
  (a single `geom_pointrange()` in place of the errorbar+point pair)
  started life as exactly this kind of custom builder, and was promoted
  to a built-in option once it proved to fit the existing `config`
  shape -- a reasonable bar to check your own custom builders against
  if you're considering proposing one upstream.

## Builder metadata: tagging a builder with `er_builder_layout()`, `er_builder_fill_role()`, `er_builder_y_role()`

The quantile builder above needed nothing beyond the function itself.
Some builders, though, make a structural or aesthetic choice that the
*composition* machinery (the code that assembles/labels/legends the
finished plot in `R/er-plot-compose.R`) needs to know about *before* it
calls the builder. Since a builder is "just a function", that
information can't live in its return value -- composition needs it
in advance, to decide things like which panel to route the builder's
output into, or how to title a legend. erplots solves this by letting a
builder carry metadata as an **attribute on the function itself**, set
by one of three small wrapper functions. Each wraps a builder and
returns it back, attribute attached, so they compose naturally with
assignment:

```r
my_builder <- er_builder_layout(my_builder, "overlay")
```

### `er_builder_layout()`: which structural family a data-layer builder belongs to

The data layer (`er_plot_add_data()`) is the one layer with two
mutually exclusive structural families a builder can be slotted into:

- `"overlay"`: a single call merged directly onto the main model panel,
  at the observations' true `(exposure, response)` coordinates (what
  `er_builder_data_overlay()`, the default, does).
- `"panel"`: one or more panels stacked *below* the base plot, the way
  `er_plot_add_groups()`'s panels are (what `er_builder_data_boxjitter()`,
  the older binary-only boxplot+jitter design, does).

`er_plot_add_data()` has to decide which of two different `config`
shapes to build (`.part_overlay()` vs. `.part_data()`) *before* it can
call the builder -- so the layout can't be inferred from what the
builder returns; it has to be knowable from the builder alone.
`er_builder_layout(builder, layout)` attaches that information as an
attribute:

```{r layout-attr}
attr(er_builder_data_overlay, "er_builder_layout")
attr(er_builder_data_boxjitter, "er_builder_layout")
```

A custom data-layer builder that omits this tag fails fast, with a
message telling you what to do, rather than silently landing in the
wrong structural slot:

```{r layout-error, error = TRUE}
untagged_builder <- function(data, config, stratify, exposure, response, strata, style) {
  ggplot2::geom_point(ggplot2::aes(x = .data[[exposure$name]], y = .data[[response$name]]))
}

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_data(builder = untagged_builder) |>
  plot()
```

Tagging it with `er_builder_layout()` fixes that. Here's a complete
custom `"overlay"`-layout data builder -- a 2D density contour in place
of raw points, useful when there are enough observations that a scatter
overplots into an unreadable smear:

```{r layout-custom}
er_builder_data_density <- er_builder_layout(
  function(data, config, stratify, exposure, response, strata, style) {
    ggplot2::geom_density2d(
      data = data,
      mapping = ggplot2::aes(x = .data[[exposure$name]], y = .data[[response$name]]),
      inherit.aes = FALSE
    )
  },
  layout = "overlay"
)

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_data(builder = er_builder_data_density) |>
  plot()
```

### `er_builder_fill_role()`: what a builder's `fill` aesthetic means

On the base plot, `fill` almost always means strata (e.g. a stratified
model ribbon). `er_builder_data_hex()` -- a built-in `"overlay"`-layout
data builder for when N is too large for a legible scatter -- is the one
exception: its `fill` encodes 2D bin density (a continuous scale), not
strata, so `.polish_labels()` needs to know to title that legend "Count"
rather than the stratification variable's label. `er_builder_fill_role(builder,
"density")` records exactly that:

```{r fill-role}
attr(er_builder_data_hex, "er_builder_fill_role")

erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())) |>
  er_plot_add_data(builder = er_builder_data_hex) |>
  plot()
```

Note the legend is titled "Count", not the (nonexistent, here) strata
label -- that's `.polish_labels()` consulting the `"density"` tag. Unlike
`er_builder_layout()`, `er_builder_fill_role()` is optional: a builder
that doesn't set it is assumed to mean strata whenever it maps `fill`
at all, which is the right default for every other builder.

### `er_builder_y_role()`: what a builder's y-axis means

`er_plot_add_groups()`'s default builders (`er_builder_group_boxplot()`,
`er_builder_group_violin()`) put the *group variable itself* on the
y-axis -- one categorical row per level -- so the group variable's own
label is the right axis title. `er_builder_group_histogram()` instead
needs its y-axis free for counts, moving group levels onto facet strips
instead; `er_builder_y_role(builder, "count")` tells `.polish_labels()` to
title the y-axis "Count" rather than the group variable's label:

```{r y-role}
attr(er_builder_group_histogram, "er_builder_y_role")

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_groups(group_by = treatment, builder = er_builder_group_histogram) |>
  plot()
```

Like `er_builder_fill_role()`, this tag is optional -- a group builder
that doesn't set it keeps the old behaviour (group variable's label on
the y-axis), which is correct for `er_builder_group_boxplot()`/
`er_builder_group_violin()`.

### Why three separate helpers, not one generic "tag" function

`er_builder_layout()`, `er_builder_fill_role()`, and `er_builder_y_role()`
all follow the same "wrapper function that attaches an attribute"
pattern, but are kept as three distinct, narrowly-named functions rather
than one `er_builder_tag(builder, key, value)`: each corresponds to a
specific decision a specific piece of composition code needs to make
(routing for `er_plot_add_data()`, a legend title for
`.polish_labels()`'s fill logic, an axis title for `.polish_labels()`'s
group logic), and giving each its own name keeps `?er_partial`'s
"Writing your own builder" section groundable in concrete examples
rather than an abstract key-value scheme.

## Summary

| Helper | Applies to | Required? | What it controls |
|---|---|---|---|
| `er_builder_layout()` | Data-layer builders only | Yes -- errors if missing | `"overlay"` (merged onto the main panel) vs. `"panel"` (stacked panels below) |
| `er_builder_fill_role()` | Any builder mapping `fill` | No -- defaults to strata | Legend title for a non-strata `fill` aesthetic (e.g. `"density"`) |
| `er_builder_y_role()` | Group-layer builders only | No -- defaults to the group variable's label | y-axis title when the y-axis isn't the group variable itself (e.g. `"count"`) |

None of this machinery is needed for a builder that draws a familiar
idiom in a familiar slot -- the crossbar example above needed no tags at
all. It exists for the less common case where a builder changes
*where* its output goes, or *what* one of its aesthetics represents,
and the rest of the plot needs to be told so it can label things
correctly. See `?er_partial` for the full public-API contract, and
[the plot grammar article](design.html) for how these layers fit
together more broadly.
