---
title: "Extending erplots"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal when writing the erplots package was to provide a flexible mini-language for exposure-response plots that pharmacometricians could use to create most such plots without needing to write hundreds of lines of ggplot2 code. One problem with doing so, however, is that it is almost impossible to anticipate *every* possible use case for the package. Inevitably, there will be some cases where the specific plot that you need can't be constructed using the functions that are supplied by erplots. This could then become frustrating for the user, if they already have code that generates most of what they need using erplots, but can't quite get exactly what they need because there's one esoteric special case that the package can't handle. 

To mitigate this risk, erplots comes with an extension mechanism: you can write your own "builder" function that will generate the one part of the exposure-response plot that you need to modify, and then erplots will handle the rest. 

This article discusses that extension mechanism, by showing you how to write a "builder" function. It assumes you're already familiar with the erplots mini-language described in [the exposure-response plotting grammar article](design.html) (layers, singleton/additive semantics, stratification); and have a good sense of the overall process of building plots with the package. 

```{r setup}
library(erplots)
library(erglm)
```

## What is a builder function?

When creating an exposure-response plot with erplots, you first call `er_plot()` to create the basic data structure, and then add a series of layers using the five layer functions (`er_plot_add_model()`, `er_plot_add_summary()`, `er_plot_add_quantiles()`, `er_plot_add_data()`, `er_plot_add_groups()`) to add the features you want to see in your plot. The key thing to note is that these layer functions don't create the ggplot2 code themselves. Instead, they delegate that work to some other function that you specify via the `style` argument. That function is referred to as the "builder" for that layer. The erplots package comes with several different builders for each layer:

| Layer | Builders |
|---|------|
| Model | `er_style_model_ribbonline()` (default), `er_style_model_line()`, `er_style_model_spaghetti()` |
| Data | `er_style_data_overlay()` (default), `er_style_data_boxjitter()`, `er_style_data_hex()` |
| Quantile | `er_style_quantile_errorbar()` (default), `er_style_quantile_errorbar_vlines()`, `er_style_quantile_pointrange()`, `er_style_quantile_pointrange_vlines()` |
| Summary | `er_style_summary_pvalue()` (default), `er_style_summary_n()`, `er_style_summary_coefficients()`, `er_style_summary_gof()` |
| Groups | `er_style_group_boxplot()` (default), `er_style_group_histogram()`, `er_style_group_violin()` |

These builders cover a wide variety of different stylistic features you might want to employ in your exposure-response plots, but if the feature you want is not available, you can write your own. As long as your builder function has the expected structure, erplots will understand what it does and use it when constructing a plot.

## The builder signature

The main constraint that you need to satisfy when writing a builder function is to make sure that it takes the expected arguments, and uses those arguments in a manner that is consistent with what the erplots grammar expects. Specifically, every builder function must have this signature"

```r
function(data, config, stratify, exposure, response, strata, theme, ...)
```

When writing the function itself, you should be aware of what erplots will pass into your function within each of these slots:


| Argument | What it is |
|---|------|
| `data` | The original data frame passed to `er_plot()`, unmodified. |
| `config` | The pre-computed configuration for this specific layer. This is different for each layer, and is discussed in more detail below. |
| `stratify` | `TRUE`/`FALSE`: whether this layer should honour `stratify_by`. |
| `exposure`, `response`, `strata` | Plot-variable metadata lists (`name`, `label`, `limits`, ...) describing the exposure, response, and stratification variables declared in `er_plot()`. |
| `theme` | Shared theming state, adjustable via `er_plot_theme()`: `theme$theme_base`, `theme$draw_key`, `theme$format_percent()`, `theme$format_number()`. |
| `...` | Extra named arguments forwarded from the corresponding `er_plot_add_*()` call's own `...`. See "Passing extra arguments to a builder" below. |

## The builder return value

The function returns a geom, or a list of geoms/other objects that can
be added to a ggplot2 plot -- nothing more. 

## Passing extra arguments

Every `er_plot_add_*()` function also takes its own `...`, forwarded
unchanged to `style` when it's called at build time. Extra arguments must be named -- they're
appended positionally after the seven standard arguments, so an unnamed
one would silently bind to the wrong parameter. A builder that doesn't
need any extra arguments simply declares `...` and ignores it, as every
builder in this article does; see `?er_style`'s own "Passing extra
arguments to a builder" section (and `er_style_model_spaghetti()`'s use
of a `seed` passed this way) for a worked example.

This `...` is for the *builder* only. It's a different mechanism from
`er_plot_add_model()`'s `predict_args`, `er_plot_add_summary()`'s
`summary_args`, and `er_vpc_add_simulated()`'s `simulate_args` -- those
are spliced into the model-interface generics themselves
(`er_predict()`/`er_summary()`/`er_simulate()`), for a model-specific
argument with no slot in their fixed contract (e.g. a landmark time for
a time-to-event model). If you're writing a custom builder, you'll
never see `predict_args`/`summary_args`/`simulate_args` -- they never
reach `config` or the builder's own `...`. If you're implementing the
model interface for your own model class instead, see [Implementing the
model interface](model-interface.html), which covers that mechanism.

## Understanding the `config` argument

`config` is where a custom builder actually gets its data from, and it
is **not** the raw `data` frame -- it's whatever erplots derived from
`data`/`exposure`/`response`/`strata` before any builder ran. A custom
builder's whole job is to turn
that already-computed `config` into ggplot2 layers; it never needs to
re-bin, re-summarise, or re-fit anything itself. Concretely:

| Layer | Key `config` field | Contents |
|---|---|------|
| Model | `config$predictions` | One row per exposure grid point, with `fit_resp`, `ci_lower`, `ci_upper` (from `er_predict()`) |
| Summary | `config$summary`, `config$p_value`, `config$corner_distance` | `config$summary` is the full, raw return value of `er_summary()` (`NULL` unless a `model` was supplied) -- see [Implementing the model interface](model-interface.html) for its `p_value`/`coefficients`/`glance` contract; `config$p_value` is that same list's `p_value` field, extracted separately since `er_style_summary_pvalue()` reads it directly; `corner_distance` is a named vector of four minimum distances (`top_left`/`top_right`/`bottom_left`/`bottom_right`), computed from `data`'s raw `(exposure, response)` points rescaled onto `[0, 1]`, used to place a label away from the observed data |
| Quantile | `config$summary` | One row per exposure-quantile bin (× stratum), with `x_mid`, `y_mid`, `ci_lower`, `ci_upper`, plus label-placement columns. `config$breaks` also holds the `n + 1` quantile cutpoints themselves (from `cut_exposure_quantile()`), which `er_style_quantile_errorbar_vlines()`/`er_style_quantile_pointrange_vlines()` use to draw bin-boundary separators |
| Data | (none extra) | The builder mostly works from `data` directly, since this layer draws raw observations rather than a summary |
| Group | `config[[group_var]]$data`, `config[[group_var]]$counts` | The subset of `data` for that grouping variable, joined to per-group sample-size labels |

## A worked example

The story above is not entirely complete, but it covers enough that we can work through an example. Let's suppose the built-in `er_style_quantile_errorbar()` and `er_style_quantile_pointrange()` alternatives all feel like the wrong way to display the data summaries by exposure quantile. Instead, you would prefer to draw these binwise summaries using a `geom_crossbar()`. This isn't usually a great choice -- more on that below -- but you may have a good reason for wanting it, so you decide to write your own builder function for this. And since the quantile layer is mostly dependent on the `config$summary` field, that's the place to start. When doing so, it helps to look at what `config$summary` actually contains, and the most reliable way to do that is to ask erplots to show you: write a tiny "spy" builder that just prints `config` and draws nothing, and pass it in as the `style` for the layer you want to inspect:

```{r inspect-config}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

inspect_config <- function(data, config, stratify, exposure, response, strata, theme, ...) {
  print(config$summary)
  list()
}

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_quantiles(bins = 6, style = inspect_config) |>
  plot()
```

Looking at this object we see that it is a tibble with one row per exposure quantile (as always, noting that the placebo group will be its own distinct group). Each column contains a summary statistic or indicator variable that you might want to use in your builder function:

- `x_mid` and `y_mid` capture mean exposure and mean response associated with the exposure bin. Our new builder will need to use these.
- `ci_lower` and `ci_upper` are the confidence interval bounds. Our builder will need these too.
- `y_mid_lbl`, `y_lwr_lbl`, `y_upr_lbl`, and `y_lbl` support the built-in label geom that `er_style_quantile_errorbar()` draws alongside its error
bar. If we wanted to use labels in our plot we would need these columns, but since we plan to skip that we can ignore them.

Now that we understand what `config$summary` already provides, writing the builder itself is very straightforward. All we need is a small function that maps those columns onto `geom_crossbar()`'s aesthetics:

```{r builder}
er_style_quantile_crossbar <- function(data, 
                                       config, 
                                       stratify, 
                                       exposure, 
                                       response, 
                                       strata, 
                                       theme, 
                                       ...) {
  ggplot2::geom_crossbar(
    data = config$summary,
    mapping = ggplot2::aes(
      x = x_mid, 
      y = y_mid, 
      ymin = ci_lower, 
      ymax = ci_upper
    ),
    inherit.aes = FALSE
  )
}
```

That's all we need here. This particular builder is simple enough that we just need to specify a single geom, and erplots will handle the rest. Here's what it looks like when we use it:

```{r builder-example-plot}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles(style = er_style_quantile_crossbar) |>
  plot()
```

This is not, in general, a good way to create an exposure-response plot: in statistical graphics, the "cross bar" visual idiom is almost always used to show distributional information (i.e., it feels like a boxplot without the whiskers). Using it to display a mean and confidence interval is almost certainly going to confuse anyone looking at your plot. Even so, it's convenient to illustrate how to design a builder function.

Before moving on, there are a few things worth noting about this builder, all of which are generalisable to any layer:

- It ignores `data`, `stratify`, `exposure`, `response`, `strata`, and
  `theme` entirely -- a builder only needs to use the arguments relevant
  to what it draws. (A stratified version would need to map
  `color = strata` and use `config$summary`'s `strata` column, plus
  `theme$draw_key` for a legend key consistent with the other layers --
  see `er_style_quantile_errorbar()`'s source for a worked stratified
  example.)
- `inherit.aes = FALSE` is there because this geom supplies its own
  `data`/`mapping`, distinct from whatever's already on the base plot
  (the model curve, in this example) -- omitting it would try to inherit
  the base plot's aesthetics and fail, since those don't include
  `ymin`/`ymax`.
- No new `config` fields were needed. `er_style_quantile_pointrange()`
  (a single `geom_pointrange()` in place of the errorbar+point pair)
  started life as exactly this kind of custom builder, and was promoted
  to a built-in option once it proved to fit the existing `config`
  shape -- a reasonable bar to check your own custom builders against
  if you're considering proposing one upstream.

## Supplying builder metadata

The quantile builder above needed nothing beyond the function itself.
Some builders, though, make a structural or aesthetic choice that the
*composition* machinery (the code that assembles/labels/legends the
finished plot) needs to know about *before* it calls the builder. Since
a builder is "just a function", that
information can't live in its return value -- composition needs it
in advance, to decide things like which panel to route the builder's
output into, or how to title a legend. erplots solves this by letting a
builder carry metadata as **attributes on the function itself**, set by
a single wrapper function, `er_style_tag()`, with one optional
argument per piece of metadata (`layout`, `fill_role`, `y_role`,
`layer`, `zorder`). It
wraps a builder and returns it back, attributes attached, so it
composes naturally with assignment, and a builder that needs more than
one tag only needs one call:

```r
my_builder <- er_style_tag(my_builder, layout = "overlay")
```

### `layout`: which structural family a data-layer builder belongs to

The data layer (`er_plot_add_data()`) is the one layer with two
mutually exclusive structural families a builder can be slotted into:

- `"overlay"`: a single call merged directly onto the main model panel,
  at the observations' true `(exposure, response)` coordinates (what
  `er_style_data_overlay()`, the default, does).
- `"panel"`: one or more panels stacked *below* the base plot, the way
  `er_plot_add_groups()`'s panels are (what `er_style_data_boxjitter()`,
  the binary-only boxplot+jitter design, does).

`er_plot_add_data()` has to decide which of two different `config`
shapes to build *before* it can call the builder -- so the layout can't
be inferred from what the
builder returns; it has to be knowable from the builder alone.
`er_style_tag(style, layout = ...)` attaches that information as an
attribute, and is the one tag that's mandatory for a data-layer builder:

```{r layout-attr}
attr(er_style_data_overlay, "er_style_layout")
attr(er_style_data_boxjitter, "er_style_layout")
```

A custom data-layer builder that omits this tag fails fast, with a
message telling you what to do, rather than silently landing in the
wrong structural slot:

```{r layout-error, error = TRUE}
untagged_builder <- function(data, config, stratify, exposure, response, strata, theme, ...) {
  ggplot2::geom_point(ggplot2::aes(x = .data[[exposure$name]], y = .data[[response$name]]))
}

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_data(style = untagged_builder) |>
  plot()
```

Tagging it with `er_style_tag()` fixes that. Here's a complete
custom `"overlay"`-layout data builder -- a 2D density contour in place
of raw points, useful when there are enough observations that a scatter
overplots into an unreadable smear:

```{r layout-custom}
er_style_data_density <- er_style_tag(
  function(data, config, stratify, exposure, response, strata, theme, ...) {
    ggplot2::geom_density2d(
      data = data,
      mapping = ggplot2::aes(x = .data[[exposure$name]], y = .data[[response$name]]),
      inherit.aes = FALSE
    )
  },
  layout = "overlay"
)

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_data(style = er_style_data_density) |>
  plot()
```

### `fill_role`: what a builder's `fill` aesthetic means

On the base plot, `fill` almost always means strata (e.g. a stratified
model ribbon). `er_style_data_hex()` -- a built-in `"overlay"`-layout
data builder for when N is too large for a legible scatter -- is the one
exception: its `fill` encodes 2D bin density (a continuous scale), not
strata, so erplots needs to know to title that legend "Count" rather
than the stratification variable's label. `er_style_tag(style,
fill_role = "density")` records exactly that:

```{r fill-role}
attr(er_style_data_hex, "er_style_fill_role")

erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())) |>
  er_plot_add_data(style = er_style_data_hex) |>
  plot()
```

Note the legend is titled "Count", not the (nonexistent, here) strata
label -- that's erplots consulting the `"density"` tag. Unlike
`layout`, `fill_role` is optional: a builder
that doesn't set it is assumed to mean strata whenever it maps `fill`
at all, which is the right default for every other builder.

### `y_role`: what a builder's y-axis means

`er_plot_add_groups()`'s default builders (`er_style_group_boxplot()`,
`er_style_group_violin()`) put the *group variable itself* on the
y-axis -- one categorical row per level -- so the group variable's own
label is the right axis title. `er_style_group_histogram()` instead
needs its y-axis free for counts, moving group levels onto facet strips
instead; `er_style_tag(style, y_role = "count")` tells erplots to title
the y-axis "Count" rather than the group variable's label:

```{r y-role}
attr(er_style_group_histogram, "er_style_y_role")

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_groups(group_by = treatment, style = er_style_group_histogram) |>
  plot()
```

Like `fill_role`, this tag is optional -- a group builder
that doesn't set it keeps the default behaviour (group variable's label
on the y-axis), which is correct for `er_style_group_boxplot()`/
`er_style_group_violin()`.

### `layer`: which `er_plot_add_*()` a builder is meant for

`layout`, `fill_role`, and `y_role` all feed the *composition*
machinery -- deciding where a builder's output goes, or how to title a
legend/axis once it's drawn. `layer` is different: it's read by the
`er_plot_add_*()` functions themselves, *before* they call the builder
at all, purely to catch a builder plugged into the wrong slot. Every
built-in builder declares it -- `er_style_quantile_errorbar()` is
tagged `layer = "quantile"`, `er_style_group_violin()` is tagged
`layer = "group"`, and so on for all five layers (`"model"`,
`"summary"`, `"quantile"`, `"data"`, `"group"` -- `"summary"` is its
own layer, [er_plot_add_summary()], independent of the model layer).
Passing a builder tagged for one layer into a different layer's
`er_plot_add_*()` call errors immediately, naming both the layer the
builder was tagged for and the layer it was actually passed to:

```{r layer-error, error = TRUE}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_data(style = er_style_quantile_errorbar)
```

```{r layer-attr}
attr(er_style_quantile_errorbar, "er_style_layer")
```

Unlike `layout`, `layer` is entirely optional -- a custom builder that
omits it is simply never checked, regardless of which
`er_plot_add_*()` it's passed to. This means existing custom builders written
before `layer` existed keep working unchanged; tagging one is purely a
way to get an earlier, more specific error if it's ever passed to the
wrong place by mistake.

### `zorder`: where an overlay-layout data builder's geoms sit in the main panel

The model, summary, and quantile layers always draw in the same fixed
order relative to each other when they share the main panel. An
`"overlay"`-layout data builder's geoms, by default, draw *last* -- on
top of all three -- which is the right choice for a sparse builder like
`er_style_data_overlay()`, whose individual points should never be
hidden behind a translucent model ribbon. But a builder whose geoms
cover the *whole* panel, with no gaps for anything underneath to show
through, would completely bury the model curve and summary annotation
by drawing last in the same way. `er_style_tag(style, zorder =
"background")` tells erplots to draw that builder's geoms *first*
instead, so the model/summary/quantile layers are drawn on top of it.
`er_style_data_hex()` is tagged this way:

```{r zorder-attr}
attr(er_style_data_hex, "er_style_zorder")
```

Here's a custom `"overlay"`-layout builder that makes the same choice --
a filled 2D density contour, which (like `er_style_data_hex()`) covers
the whole panel:

```{r zorder-custom}
mod_gaussian <- erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())

er_style_data_density_fill <- er_style_tag(
  function(data, config, stratify, exposure, response, strata, theme, ...) {
    ggplot2::geom_density_2d_filled(
      data = data,
      mapping = ggplot2::aes(x = .data[[exposure$name]], y = .data[[response$name]]),
      contour_var = "ndensity",
      show.legend = FALSE
    )
  },
  layout = "overlay",
  zorder = "background"
)

erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(mod_gaussian) |>
  er_plot_add_data(style = er_style_data_density_fill) |>
  plot()
```

The model curve stays visible because the filled contours were drawn
first. Omitting the `zorder` tag (or setting it to `"foreground"`,
equivalent to the default) draws exactly the same contours *after* the
model curve instead, burying it completely:

```{r zorder-foreground}
er_style_data_density_fill_fg <- er_style_tag(
  function(data, config, stratify, exposure, response, strata, theme, ...) {
    ggplot2::geom_density_2d_filled(
      data = data,
      mapping = ggplot2::aes(x = .data[[exposure$name]], y = .data[[response$name]]),
      contour_var = "ndensity",
      show.legend = FALSE
    )
  },
  layout = "overlay"
)

erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(mod_gaussian) |>
  er_plot_add_data(style = er_style_data_density_fill_fg) |>
  plot()
```

`zorder` only has an effect on an `"overlay"`-layout data builder. A
`"panel"`-layout builder's geoms (e.g. `er_style_data_boxjitter()`) are
drawn in their own, separate patchwork panel, never sharing space with
the model/summary/quantile layers, so the tag is inert there if set.

### One function, five independent arguments

`layout`, `fill_role`, `y_role`, `layer`, and `zorder` are all set via
the same `er_style_tag()` call rather than five separate wrapper
functions. Each argument is independent and optional (aside from
`layout` being mandatory for a data-layer builder specifically -- see
above), so a builder that needs to declare more than one piece of
metadata -- say, a custom "overlay"-layout data builder whose `fill`
also means something other than strata, and whose geoms cover the
whole panel -- can do it in one call:

```r
my_density_builder <- er_style_tag(
  my_density_builder,
  layout = "overlay",
  fill_role = "density",
  layer = "data",
  zorder = "background"
)
```

This is close to what the built-in `er_style_data_hex()` does (it
sets `layout`, `fill_role`, `layer`, and `zorder` together).

### Summary

| Argument | Applies to | Required? | What it controls |
|---|---|---|---|
| `layout` | Data-layer builders only | Yes -- errors if missing | `"overlay"` (merged onto the main panel) vs. `"panel"` (stacked panels below) |
| `fill_role` | Any builder mapping `fill` | No -- defaults to strata | Legend title for a non-strata `fill` aesthetic (e.g. `"density"`) |
| `y_role` | Group-layer builders only | No -- defaults to the group variable's label | y-axis title when the y-axis isn't the group variable itself (e.g. `"count"`) |
| `layer` | Any builder | No -- unchecked if unset | Which `er_plot_add_*()` the builder is meant for; mismatches error immediately |
| `zorder` | `"overlay"`-layout data builders only | No -- defaults to `"foreground"` | `"foreground"` (drawn after model/summary/quantile) vs. `"background"` (drawn before) |

None of this machinery is needed for a builder that draws a familiar
idiom in a familiar slot -- the crossbar example above needed no tags at
all. It exists for the less common case where a builder changes
*where* its output goes, or *what* one of its aesthetics represents,
and the rest of the plot needs to be told so it can label things
correctly. See `?er_style` for the full public-API contract, and
[the exposure-response plotting grammar article](design.html) for how these layers fit
together more broadly.

## Writing a custom builder for `er_vpc()`

Everything above is about `er_plot()`'s own builders. `er_vpc()`
(see [Visual predictive checks](plot-vpc.html)) has its own, narrower
shared signature:

```r
function(data, config, exposure, response, theme, ...)
```

Two arguments from `er_plot()`'s signature are missing: `stratify` and
`strata`. A VPC builder never needs to know about stratification
directly -- an optional `stratify_by` only ever facets the finished plot
(`ggplot2::facet_wrap()`), never recolours it, and every `config` table a
builder reads from already carries the `.vpc_stratum` column that
facet needs. A builder just draws its geoms once, the same way whether
the object is stratified or not; faceting happens afterwards, outside
the builder entirely.

`config$summary` is the table both `er_style_vpc_observed_mean_errorbar()`
and `er_style_vpc_simulated_mean_errorbar()` (the defaults) plot from: one
row per bin, with `y_mid`/`ci_lower`/`ci_upper` and, depending on
`config$is_numeric_group`, either `.vpc_bin` (a categorical `plot_by`) or
`x_median`/`x_mid` (a numeric one) to plot at. `config$percentiles` is the
table the quantile-line/quantile-errorbar idioms read instead -- one row
per bin *and* per requested percentile -- and is only computed for a
continuous/count response with `probs` set; a builder that expects it
should check for `NULL` the way every built-in one does, or declare
`response_types` (below) so `er_vpc_add_observed()`/`er_vpc_add_simulated()`
catch the mismatch earlier.

### A worked example

Suppose the default point-and-solid-errorbar look for the observed layer
doesn't stand out enough against a ribbon-heavy simulated layer. Diamond
markers with a dashed error bar are one easy way to make the observed
side visually distinct, built the same way `er_style_vpc_observed_mean_errorbar()`
itself is, straight from `config$summary`:

```{r vpc-custom-builder}
er_style_vpc_observed_diamond <- er_style_tag(
  function(data, config, exposure, response, theme, ...) {
    x_var <- if (config$is_numeric_group) "x_median" else ".vpc_bin"
    list(
      ggplot2::geom_errorbar(
        data = config$summary,
        mapping = ggplot2::aes(x = .data[[x_var]], ymin = ci_lower, ymax = ci_upper, color = "Observed"),
        width = 0.15, linetype = "dashed",
        inherit.aes = FALSE
      ),
      ggplot2::geom_point(
        data = config$summary,
        mapping = ggplot2::aes(x = .data[[x_var]], y = y_mid, color = "Observed"),
        shape = 18, size = 3,
        inherit.aes = FALSE
      )
    )
  },
  layer = "observed",
  response_types = c("binary", "continuous", "count"),
  plot_by_types = c("continuous", "discrete"),
  marker_source = "summary"
)
```

Plugged in for the observed layer only -- there's no requirement to
customise both sides of the comparison at once -- it pairs without
incident against the default simulated builder:

```{r vpc-custom-plot}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |>
  er_vpc(exposure = aucss, response = ae1) |>
  er_vpc_add_observed(style = er_style_vpc_observed_diamond) |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
```

### VPC-specific `er_style_tag()` arguments

`er_vpc_add_observed()`/`er_vpc_add_simulated()` recognise four
[er_style_tag()] arguments beyond `layer` (checked the same way as
`er_plot_add_*()`'s own `layer` tag above), none of which apply to an
`er_plot()` builder:

| Argument | Applies to | Required? | What it controls |
|---|---|---|---|
| `response_types` | Any VPC builder | No -- unchecked if unset | Which of `"binary"`/`"continuous"`/`"count"` responses the builder supports |
| `plot_by_types` | Any VPC builder | No -- unchecked if unset | Whether the builder supports a `"continuous"` (numeric) `plot_by`, a `"discrete"` (categorical) one, or both |
| `layout` | Any VPC builder | No -- unchecked if unset | `"categorical"` (plots at discrete bin locations) vs. `"continuous"` (plots at each bin's numeric midpoint) -- **different values from the data layer's own `"overlay"`/`"panel"` pair above**, and checked between the observed and simulated builder rather than against a fixed rule |
| `marker_source` | Any VPC builder | No -- unchecked if unset | Which config table (`"summary"` or `"percentiles"`) the builder actually draws its marker(s) from, read by `er_vpc_theme()`'s `xlim`/`ylim` clipping check (see [Visual predictive checks](plot-vpc.html#theming)) |

`response_types`/`plot_by_types` are checked against the VPC object's own
data, not against another builder -- `er_style_vpc_observed_quantile_line()`
declares `response_types = c("continuous", "count")` (it needs
`config$percentiles`, never computed for a binary response), so pairing
it with a binary-response VPC errors immediately, before any binning
happens:

```{r vpc-response-type-error, error = TRUE}
erglm_data |>
  er_vpc(exposure = aucss, response = ae1) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line)
```

`layout`, by contrast, is checked between the *two* builders passed to
`er_vpc_add_observed()`/`er_vpc_add_simulated()`, not against the data --
pairing a `"categorical"`-layout builder with a `"continuous"`-layout one
would otherwise plot the two sides of the comparison at inconsistent
x-positions for the same bin:

```{r vpc-layout-error, error = TRUE}
stub_categorical <- er_style_tag(
  function(data, config, exposure, response, theme, ...) list(),
  layer = "observed", layout = "categorical"
)
stub_continuous <- er_style_tag(
  function(data, config, exposure, response, theme, ...) list(),
  layer = "simulated", layout = "continuous"
)

mod_gauss <- erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())

erglm_data |>
  er_vpc(exposure = aucss, response = biomarker_change) |>
  er_vpc_add_observed(style = stub_categorical) |>
  er_vpc_add_simulated(model = mod_gauss, seed = 1234, style = stub_continuous)
```

None of the built-in idioms actually disagree this way -- the two
built-in `layout`-tagged builders (`er_style_vpc_observed_quantile_line()`/
`er_style_vpc_simulated_quantile_ribbon()`) are already a matched pair,
and every other built-in pair leaves `layout` untagged entirely, since
their x-position adapts to `plot_by`'s type at build time rather than
being fixed. The check exists for a custom builder that *does* commit to
one family or the other.

## Writing a custom builder for `er_tte()`

`er_tte()` (see [Time-to-event plots](plot-tte.html)) has its own shared
signature too, closer to `er_plot()`'s than `er_vpc()`'s:

```r
function(data, config, stratify, time, strata, theme, ...)
```

`time` replaces `exposure`/`response` -- a time-to-event plot has one
axis-pair variable, not two -- and otherwise the shape matches
`er_plot()`'s exactly, including the same `stratify`/`strata` pair a VPC
builder doesn't get, since TTE stratification *does* map to colour/fill
directly (one Kaplan-Meier curve per stratum, same panel), unlike VPC's
facet-only stratification.

`config`'s contents differ by layer: the curve layer gets `config$table`
(the tidy Kaplan-Meier table, with a `(0, 1)` origin row already
prepended) and `config$time_upper`; the censor layer gets a `config$table`
already filtered to censoring events; the risktable layer gets its own
`config$table`/`config$breaks`; the pvalue layer gets `config$p_value`/
`config$corner_distance`; the model layer gets `config$predictions`/
`config$time_grid`. A custom builder for a given layer reads whichever of
these its own layer actually populates -- see each layer's own
`er_style_tte_*()` help page for the exact shape.

### A worked example

A Kaplan-Meier confidence band is drawn as a series of `ggplot2::geom_rect()`
rectangles (see `er_style_tte_curve_km()`'s own documentation for why),
which makes building one from scratch more work than it's worth for a
small addition like a reference line at 50% survival. Wrapping the
built-in curve builder and adding one more geom is the more direct route:

```{r tte-custom-builder}
library(survival)

er_style_tte_curve_median <- er_style_tag(
  function(data, config, stratify, time, strata, theme, ...,
           show_ci = TRUE, ribbon_alpha = 0.15, linewidth = 1) {
    geoms <- er_style_tte_curve_km(
      data, config, stratify, time, strata, theme,
      show_ci = show_ci, ribbon_alpha = ribbon_alpha, linewidth = linewidth
    )
    c(geoms, list(
      ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40")
    ))
  },
  layer = "curve"
)
```

```{r tte-custom-plot}
lung_sex <- lung |> transform(sex = factor(sex, labels = c("Male", "Female")))

lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve(style = er_style_tte_curve_median) |>
  plot()
```

The `layer` tag works identically to `er_plot()`'s own -- the same
`er_style_tag(fn, layer = ...)` mechanism, checked by every
`er_tte_add_*()` function against the layer it was actually called from,
just with five more valid values (`"curve"`, `"censor"`, `"risktable"`,
`"pvalue"`, alongside `"model"` for `er_tte_add_model()`). A builder
tagged for the wrong one errors immediately:

```{r tte-layer-error, error = TRUE}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve(style = er_style_tte_censor_ticks)
```

None of `er_vpc()`'s own extra tags (`response_types`, `plot_by_types`,
the VPC flavour of `layout`, `marker_source`) apply to a TTE builder --
`er_tte()` has only one response variable shape (a `Surv()`-style
time/event pair) and one x-axis variable (`time`, always continuous), so
there's nothing for those tags to distinguish.

See `?er_style_tte_curve`/`?er_style_tte_censor`/`?er_style_tte_risktable`/
`?er_style_tte_pvalue`/`?er_style_tte_model` for each layer's own
`config` contents in full, and
[The exposure-response plotting grammar](design.html)/[Implementing the model
interface](model-interface.html) for how `er_tte_add_model()`'s builder
relates to [er_predict_survival()] specifically.
