---
title: "The built plot object (internal structure)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **This article documents an implementation detail, not a public interface.**
> Everything described below -- the shape of `object$plot`, `object$layer`,
> and `object$output`, the names used internally, and the internal
> `.polish_*()`/`.build_*()` helpers -- is unexported, undocumented in
> `?er_plot`, and not covered by erplots' usual
> conventions. It can, and likely will, change
> in a future release without notice or a deprecation period. Nothing in
> this article should be relied on in package code, and any script that
> depends on it should be treated as tied to the erplots version it was
> written against.
>
> The preferred way to change how a plot looks is a custom `style` function
> (see [Extending erplots](extending.html)). Read this article only if
> you've hit a case that mechanism genuinely can't reach -- e.g. a one-off
> tweak to a specific panel's theme, or a fix you need *right now* and don't
> want to write and register a whole custom builder for -- and you're
> willing to reach into the object and patch it by hand.

```{r setup}
library(erplots)
library(erglm)

mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

p <- erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles() |>
  er_plot_add_data(style = er_style_data_boxjitter)
```

## Two stages: `object$plot`, then `object$output`

Calling `plot()` (or `print()`) on an `er_plot` object calls `er_plot_build()`
first. `er_plot_build()` does its work in two stages, and both intermediate
results are kept on the returned object rather than discarded:

```{r build}
built <- er_plot_build(p)
names(built)
```

**Stage one** builds one ordinary ggplot2 object per layer family, stored
under `built$plot`. **Stage two** ("polishing": margins, axis/legend labels,
discrete colour scales, panel arrangement, legend deduplication, and the
plot-level theme) turns those separate ggplot2 objects into a single
[patchwork](https://patchwork.data-imaginist.com/) object, stored under
`built$output`. `plot.er_plot()` itself is little more than

```r
plot.er_plot <- function(x, y = NULL, ...) {
  object <- er_plot_build(x)
  plot(object$output)
}
```

so `built$output` is exactly what you see rendered.

## `object$plot`: one ggplot2 object per layer family

`built$plot` is a plain list with up to three slots -- `base`, `data`,
`group` -- each either `NULL` or containing real ggplot2 object(s):

```{r plot-names}
names(built$plot)
class(built$plot$base)
```

- **`base`** is a single ggplot2 object. It carries the model curve/ribbon,
  the quantile summary geoms, and the summary annotation -- whichever of
  those layers are present -- and only exists at all if at least one of
  them is (or if the plot has no layers whatsoever, in which case it's an
  empty axes-only canvas). An `"overlay"`-layout data builder (the default,
  `er_style_data_overlay()`) also draws directly onto `base`, rather than
  into its own panel -- so a plot built with the default data style has a
  `NULL` `built$plot$data` even though it clearly *has* a data layer:

  ```{r overlay-example}
  p_overlay <- erglm_data |> er_plot(aucss, ae1) |> er_plot_add_data()
  built_overlay <- er_plot_build(p_overlay)
  names(built_overlay$plot)
  is.null(built_overlay$plot$data)
  ```

- **`data`** is a *named list* of ggplot2 objects -- never a single bare
  object, even when there's only one panel -- used only by a
  `"panel"`-layout data builder (e.g. `er_style_data_boxjitter()`, used in
  `p` above). The names are meaningful and depend on response type and
  stratification: `"upper"`/`"lower"` for the binary responder/non-responder
  split, `"data"` for a single unstratified continuous/count panel, or one
  name per stratum level when stratified and faceted.

  ```{r data-panels}
  names(built$plot$data)
  ```

- **`group`** is a named list of ggplot2 objects, one per grouping variable
  added via `er_plot_add_groups()`, keyed by variable name. `NULL` if no
  group layer was added.

## `object$output`: the final patchwork object

```{r output-class}
class(built$output)
```

`built$output` is what `patchwork::wrap_plots()` produces from the
individual panels in `built$plot` (stacked vertically via `ncol = 1`, sized
by `er_plot_theme()`'s `height_*` arguments, with `guides = "collect"` and
`axes = "collect"` to merge legends and align axes across panels), plus
`patchwork::plot_annotation()` for the plot-level title/subtitle/caption.

A patchwork object behaves like an ordinary list of its constituent plots
for indexing purposes, which is the most direct route to a hand patch: pull
out the panel you want, modify it like any ggplot2 object, and put it back.

```{r manual-patch}
length(built$output)
built$output[[1]] <- built$output[[1]] + ggplot2::labs(title = "Responders")
built$output
```

This is usually less error-prone than editing `built$plot` and re-running
the polishing steps yourself, since it works entirely with ordinary,
already-composed ggplot2 objects and doesn't need to know about the
internal `.polish_*()` helpers at all. Its limitation is exactly that it
operates *after* polishing -- e.g. adding a new discrete colour mapping this
way won't get the cross-panel legend deduplication that
`er_plot_theme(color_discrete = ...)` gets, because that logic already ran.

## Summary

| Object | Type | Contains |
|---|---|---|
| `built$plot$base` | single ggplot object, or `NULL` | model/summary/quantile geoms, and an `"overlay"`-layout data builder's points |
| `built$plot$data` | named list, or `NULL` | one panel per `"panel"`-layout data builder's output |
| `built$plot$group` | named list, or `NULL` | one panel per `er_plot_add_groups()` variable |
| `built$output` | single `patchwork` object | all of the above, stacked, themed, and annotated |

Treat all of this as a snapshot of the current implementation, not a
contract -- see the warning at the top of this article.
