---
title: "Implementing the model interface"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

erplots never fits a model. Every plot it draws is built from predictions,
simulations, or summary statistics that some *other* object hands it, via
three S3 generics: `er_predict()`, `er_simulate()`, and `er_summary()`.
Collectively these are "the model interface", documented tersely at
`?er_model_interface`. This article exists because that help topic is a
contract, not a tutorial: it tells you what each generic must return, but
not what implementing one actually looks like, or why the contract is
shaped the way it is. If you maintain a modelling package and want its
model objects to work with `er_plot_add_model()` and friends, this is the
article to read.

You're assumed to already know your way around S3 dispatch, `predict()`
methods, and the general shape of a modelling package -- this article
doesn't re-explain those. What it does explain is the erplots-specific
part: which three generics to implement, what each one is *for*, and what
erplots does with the result once you've implemented it.

```{r setup}
library(erplots)
library(erglm)
```

## Why a generic interface at all?

Exposure-response plots need model predictions, but "model predictions"
means something different depending on what fit the model: a `glm`'s
`predict(type = "response")`, an `nls` object's own delta-method
standard errors, a Bayesian model's posterior draws. erplots doesn't want
to know about any of that -- it wants one stable shape it can build a
`geom_ribbon()` from, regardless of what produced it. `er_predict()` is
that shape. The same reasoning extends to `er_simulate()` (uncertainty
visualisation) and `er_summary()` (annotation text): each is a narrow,
specific question erplots needs answered, phrased so that *any* model
object can answer it by implementing one S3 method.

Only `er_predict()` is required. `er_simulate()` and `er_summary()` are
opt-in: their default methods return `NULL`, which every part of erplots
that calls them treats as "not available" rather than an error. A model
package can implement just `er_predict()` and get the model curve/ribbon
layer working; adding `er_simulate()` unlocks spaghetti plots and VPCs;
adding `er_summary()` unlocks p-value/coefficient/goodness-of-fit
annotations. None of the three depends on any other being implemented.

## `er_predict()`: the one you must write

```r
er_predict(model, newdata, conf_level = 0.95, ...)
```

Given a fitted `model` and a `newdata` data frame of covariate values,
return `newdata` with three columns added: `fit_resp` (the point
prediction, on the response scale), `ci_lower`, and `ci_upper` (a
`conf_level` confidence interval around it, also on the response scale).
Nothing else is required -- no particular column order, no class on the
return value, no restriction on what other columns `newdata` may already
have (erplots doesn't touch them).

Two things are worth calling out explicitly because they're easy to get
wrong:

- **The interval is on the response scale, not the link scale.** For a
  binary-response GLM this matters: computing the interval on the linear
  predictor and then back-transforming with the inverse link gives an
  asymmetric, properly-bounded-in-`[0, 1]` interval; computing it directly
  on the fitted probabilities does not. Get this right at the source,
  since erplots has no way to tell (or fix) an interval that was computed
  on the wrong scale.
- **`newdata` is *returned*, not just used.** `er_predict()`'s contract is
  "augment and return `newdata`", not "return a new three-column data
  frame". This matters when `newdata` carries a stratification column --
  erplots builds `newdata` as an exposure grid crossed with every stratum
  level when the plot is stratified, and expects that column to survive
  the round trip through `er_predict()` unchanged.

### Arguments beyond `model`/`newdata`/`conf_level`

The fixed signature above covers every model this article's two toy
classes need, but some models need something else entirely -- a
landmark time for a model that reduces a survival curve to `P(event by
t*)`, say, with nowhere else in `er_predict(model, newdata, conf_level,
...)`'s contract for that value to live. That's what the trailing `...`
in the generic's own signature is for: implement your method as
`er_predict.your_class(model, newdata, conf_level = 0.95, landmark_time
= NULL, ...)`, and a caller reaches it via `er_plot_add_model()`'s
`predict_args` argument, kept deliberately separate from that same
function's own `...` (which reaches the `style` builder instead --
see [Extending erplots](extending.html)):

```r
er_plot_add_model(your_model, predict_args = list(landmark_time = 90))
```

### A worked example

Suppose you're writing a small modelling package of your own. You fit
models with a thin wrapper around `stats::glm()`, and want those model
objects to work with erplots. Tag the class first, so dispatch has
something to catch:

```{r toy-model-fit}
fit_toy_model <- function(formula, data, family = gaussian()) {
  fit <- glm(formula, data = data, family = family)
  class(fit) <- c("toy_model", class(fit))
  fit
}

mod <- fit_toy_model(ae1 ~ aucss, erglm_data, family = binomial())
class(mod)
```

Prepending `"toy_model"` (rather than replacing the class vector) keeps
every existing `glm`/`lm` method -- `predict()`, `summary()`, `coef()`,
`vcov()` -- working via ordinary S3 inheritance; `er_predict.toy_model()`
only needs to add the three columns erplots wants, using
`stats::predict.glm()` to do the actual work:

```{r toy-predict}
er_predict.toy_model <- function(model, newdata, conf_level = 0.95, ...) {
  z <- -qnorm((1 - conf_level) / 2)
  link <- predict(model, newdata = newdata, se.fit = TRUE, type = "link")
  linkinv <- family(model)$linkinv

  newdata$fit_resp <- linkinv(link$fit)
  newdata$ci_lower <- linkinv(link$fit - z * link$se.fit)
  newdata$ci_upper <- linkinv(link$fit + z * link$se.fit)
  newdata
}
```

That's the whole method. It's already enough to plug into the mini
language:

```{r toy-predict-plot}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  plot()
```

Not every model has a `predict()` method that computes standard errors
for you, though. An `nls` fit doesn't -- `predict.nls()` silently ignores
`se.fit`. When there's no ready-made standard error, the delta method
(a numerical gradient of the fitted-value function with respect to the
parameters, combined with the parameter covariance matrix) is a
general-purpose fallback that works for any differentiable mean function:

```{r toy-emax-predict}
emax_fun <- function(par, aucss) {
  par[["e0"]] + par[["emax"]] * aucss / (par[["ec50"]] + aucss)
}

er_predict.toy_emax <- function(model, newdata, conf_level = 0.95, ...) {
  z <- -qnorm((1 - conf_level) / 2)
  par <- coef(model)
  fit <- emax_fun(par, newdata$aucss)

  eps <- 1e-6
  grad <- vapply(names(par), function(nm) {
    par_up <- par; par_up[nm] <- par_up[nm] + eps
    par_dn <- par; par_dn[nm] <- par_dn[nm] - eps
    (emax_fun(par_up, newdata$aucss) - emax_fun(par_dn, newdata$aucss)) / (2 * eps)
  }, FUN.VALUE = numeric(nrow(newdata)))
  se <- sqrt(rowSums((grad %*% vcov(model)) * grad))

  newdata$fit_resp <- fit
  newdata$ci_lower <- fit - z * se
  newdata$ci_upper <- fit + z * se
  newdata
}

emax_mod <- nls(
  biomarker_change ~ e0 + emax * aucss / (ec50 + aucss),
  data = erglm_data,
  start = list(e0 = -2, emax = 5, ec50 = 1000)
)
class(emax_mod) <- c("toy_emax", class(emax_mod))

erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(emax_mod) |>
  plot()
```

This `emax_fun()`/finite-difference pairing is deliberately hand-rolled,
to keep this article dependency-free; a real implementation would more
likely reach for `numDeriv::jacobian()`, or -- if the fitting function
already has an analytic gradient available (as `nls()` itself does,
internally) -- use that instead of finite differences.

### A worked example: an argument beyond the fixed contract

To see `predict_args` actually reach a method, extend `toy_model` with
a landmark-style transform: a (deliberately simplistic) rescaling of
the fitted probability by how far a caller-supplied `landmark_time` is
from a fixed reference. `landmark_time` has no slot in `er_predict()`'s
fixed `model`/`newdata`/`conf_level` contract, so it can only arrive
through `...`:

```{r toy-landmark-predict}
er_predict.toy_landmark <- function(model, newdata, conf_level = 0.95, landmark_time = NULL, ...) {
  if (is.null(landmark_time)) {
    rlang::abort("`er_predict.toy_landmark()` requires `landmark_time`.")
  }
  out <- er_predict.toy_model(model, newdata, conf_level = conf_level)
  scale <- landmark_time / 90
  out$fit_resp <- pmin(out$fit_resp * scale, 1)
  out$ci_lower <- pmin(out$ci_lower * scale, 1)
  out$ci_upper <- pmin(out$ci_upper * scale, 1)
  out
}

landmark_mod <- mod
class(landmark_mod) <- c("toy_landmark", class(landmark_mod))
```

Omitting `landmark_time` errors, rather than silently plotting an
unscaled curve:

```{r toy-landmark-error, error = TRUE}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(landmark_mod) |>
  plot()
```

Supplying it via `predict_args` reaches the method -- and, since
`landmark_time` genuinely changes what's plotted (a shorter landmark
means a lower probability of the event having happened by then), a
smaller value visibly flattens the curve:

```{r toy-landmark-plot}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(landmark_mod, predict_args = list(landmark_time = 30)) |>
  plot()
```

The same `predict_args` list happily coexists with a `style` builder
that has its own, unrelated extra arguments via `...` -- the two never
collide, because they're spliced into two different calls
(`er_predict()` and `style`, respectively):

```{r toy-landmark-plot-spaghetti}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(
    landmark_mod,
    style = er_style_model_spaghetti,
    predict_args = list(landmark_time = 90),
    seed = 8213
  ) |>
  plot()
```

`er_plot_add_summary()`'s `summary_args` and `er_vpc_add_simulated()`'s
`simulate_args` work the same way, for `er_summary()`/`er_simulate()`
methods that need an argument beyond what's covered below.

## `er_simulate()`: for spaghetti plots and VPCs

```r
er_simulate(model, newdata, nsim = 100, seed = NULL, ...)
```

Return a data frame of `nsim` replicates of `newdata`, stacked
row-wise, with a `sim_id` column identifying which replicate each row
belongs to, and a `fit_resp` column giving that replicate's simulated
prediction. Don't implement this generic at all if your model can't
support it -- the default method returns `NULL`, and every caller
(`er_style_model_spaghetti()`, `er_vpc_add_simulated()`) checks for
`NULL` and treats it as "simulation isn't available for this model",
not an error.

The one subtlety worth understanding is that `fit_resp` here means
something narrower than it sounds like. It's a draw from the *parameter
uncertainty* in the model -- "if the true parameters were slightly
different (as they plausibly could be, given how much data you fit on),
where would the mean curve sit?" -- not a draw of an actual observation.
That's enough for a spaghetti plot, where every faint line is one
plausible mean curve. It is *not* enough for a visual predictive check,
which needs to compare simulated *observations* (with all their sampling
noise) against observed ones -- a materially different, noisier
quantity. `er_simulate()`'s contract has an independently optional second
column for exactly this, `sim_resp`: a full response-scale draw that adds
observation-level noise on top of the parameter draw (a 0/1 draw for a
binary response, an integer draw for a count response, a draw that
includes residual variance for a continuous response). A method can
return `fit_resp` alone, or both columns from the same call; there's no
reason to compute `sim_resp` if you only care about spaghetti plots, but
`er_vpc_add_simulated(model = ...)` will refuse to run (with an
informative error, not a silently-too-narrow VPC band) if `sim_resp` is
missing.

### A worked example

Continuing the `toy_model` class from above -- drawing parameter vectors
from a multivariate normal approximation to the sampling distribution of
`coef(model)` is the standard move for a GLM, and gives you `fit_resp`
for free once you have the draws:

```{r toy-simulate}
er_simulate.toy_model <- function(model, newdata, nsim = 100, seed = NULL, ...) {
  if (!is.null(seed)) set.seed(seed)

  beta_hat <- coef(model)
  beta_draws <- mvtnorm::rmvnorm(nsim, mean = beta_hat, sigma = vcov(model))
  linkinv <- family(model)$linkinv
  X <- model.matrix(delete.response(terms(model)), data = newdata)

  reps <- vector("list", nsim)
  for (i in seq_len(nsim)) {
    eta <- as.vector(X %*% beta_draws[i, ])
    fit_resp <- linkinv(eta)
    reps[[i]] <- newdata |>
      dplyr::mutate(
        sim_id = i,
        fit_resp = fit_resp,
        # the extra ingredient `sim_resp` needs beyond `fit_resp`:
        # observation-level sampling noise, appropriate to the response
        # type (here, a binary draw at the simulated probability)
        sim_resp = rbinom(length(fit_resp), size = 1, prob = fit_resp)
      )
  }
  dplyr::bind_rows(reps)
}
```

With this in place, both the spaghetti plot and the VPC work:

```{r toy-simulate-plot}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod, style = er_style_model_spaghetti, seed = 8213) |>
  plot()

erglm_data |>
  er_vpc(aucss, ae1) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, nsim = 50, seed = 4471) |>
  plot()
```

Had `er_simulate.toy_model()` only ever populated `fit_resp`, the
spaghetti plot above would still work unchanged, but the
`er_vpc_add_simulated()` call would error, naming `sim_resp` as the
missing ingredient, rather than quietly drawing a too-narrow band from
`fit_resp` alone.

The `seed` handling above is a plain `set.seed()` for readability;
erglm's own `er_simulate.erglm_model()` additionally auto-picks and
*reports* a seed when the caller doesn't supply one (via
`rlang::inform()`), so a plot is reproducible even when nobody thought to
pass `seed = ` -- worth considering for your own implementation if
reproducibility of un-seeded calls matters to your users.

Like `er_predict()`, `er_simulate()` may need an argument beyond
`model`/`newdata`/`nsim`/`seed` -- a `toy_landmark`-style
`er_simulate.your_class(model, newdata, nsim, seed, landmark_time =
NULL, ...)`, say. `er_vpc_add_simulated()`'s `simulate_args` reaches it
the same way `er_plot_add_model()`'s `predict_args` reaches
`er_predict()` above:

```r
er_vpc_add_simulated(vpc, model = your_model, simulate_args = list(landmark_time = 90))
```

## `er_summary()`: annotation text

```r
er_summary(model, ...)
```

Return `NULL` (again, the default -- don't implement this generic if
there's nothing sensible to report) or a named list with any of three
independently-optional keys: `p_value`, `coefficients`, `glance`.
Unrecognized keys are permitted and ignored by the built-in summary
builders, which gives your own package room to stash extra fields for a
custom builder to read, without erplots ever needing to know about them.
`er_plot_add_summary()` always forwards `conf_level` to `er_summary()`
(useful for a `coefficients` result's `conf_low`/`conf_high` columns --
neither toy method below needs it, but `erglm::er_summary.erglm_model()`
does), plus anything supplied via its `summary_args` argument -- the
same mechanism as `er_predict()`'s `predict_args`, for an argument
beyond `model`/`conf_level`.

- **`p_value`** -- a single headline p-value (or `NULL`), for a model
  with one unambiguous exposure effect. `er_style_summary_pvalue()`
  reads this directly. Only return a non-`NULL` value here when there
  really is one privileged term; picking one arbitrarily out of several
  candidates would be misleading.
- **`coefficients`** -- a tibble/data frame, one row per model parameter,
  for models where no single term deserves top billing. Required
  columns: `term`, `estimate`. Optional: `label` (a display name,
  falling back to `term` if absent), `std_error`, `statistic`,
  `p_value`, `conf_low`, `conf_high` (each `NA` where not
  computed/meaningful for that row). Read by
  `er_style_summary_coefficients()`. Column names are snake_case, not
  `broom::tidy()`'s dotted names -- matching this package's existing
  convention elsewhere (`p_value`, `corner_distance`).
- **`glance`** -- a single-row tibble/data frame of model-level
  goodness-of-fit, `broom::glance()`-style: optional `n`, `df_residual`,
  `logLik`, `aic`, `bic`, `deviance`, `r_squared` (`NA` if not
  meaningful -- e.g. for a non-Gaussian model), `converged`. Read by
  `er_style_summary_gof()`.

These three keys are independent, not mutually exclusive alternatives: a
model with one privileged coefficient can (and, for erglm, does) return
all three at once -- `p_value` for a quick annotation,
`coefficients`/`glance` for anyone who wants more detail via a different
summary builder. A model with no privileged term, on the other hand,
should return `p_value = NULL` explicitly rather than omit the key, so
callers can tell "there genuinely isn't one" apart from "this method
hasn't been updated yet".

### A worked example: the single-coefficient case

For `toy_model`, `aucss`'s coefficient is the one term worth headlining:

```{r toy-summary}
er_summary.toy_model <- function(model, ...) {
  coefs <- summary(model)$coefficients
  if (nrow(coefs) < 2) return(NULL)  # intercept only, nothing to report
  list(p_value = unname(coefs[2, "Pr(>|z|)"]))
}

erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_summary(model = mod) |>
  plot()
```

### A worked example: the multi-parameter case

`toy_emax`'s three parameters (`e0`, `emax`, `ec50`) have no single
"the" effect -- there's no term you could headline with a lone p-value
without implying the other two don't matter. This is exactly the
`coefficients`-instead-of-`p_value` case the contract exists for:

```{r toy-emax-summary}
er_summary.toy_emax <- function(model, ...) {
  s <- summary(model)$coefficients
  coefficients <- tibble::tibble(
    term = rownames(s),
    estimate = s[, "Estimate"],
    std_error = s[, "Std. Error"],
    statistic = s[, "t value"],
    p_value = s[, "Pr(>|t|)"]
  )
  list(p_value = NULL, coefficients = coefficients)
}

erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(emax_mod) |>
  er_plot_add_summary(model = emax_mod, style = er_style_summary_coefficients) |>
  plot()
```

Note the explicit `p_value = NULL` above, and the explicit `style =
er_style_summary_coefficients` on the `er_plot_add_summary()` call --
`er_style_summary_pvalue()` is still the *default* summary builder
regardless of which model is attached, and it would silently draw
nothing here (correctly: it reads `config$p_value`, sees `NULL`, and
has nothing to show) rather than falling back to `coefficients` on your
behalf. Matching a model whose `er_summary()` populates `coefficients`
with a builder that actually reads `coefficients` is left to you, the
caller -- `er_summary()`'s job is only to make the data available.

## Testing your implementation

There's no formal conformance checker for this interface -- it's three
plain S3 generics, so the real test is simply exercising the layers that
call them:

- `er_plot_add_model(your_model)` exercises `er_predict()` alone.
- `er_plot_add_model(your_model, style = er_style_model_spaghetti)`
  additionally exercises `er_simulate()`'s `fit_resp`.
- `er_vpc_add_simulated(model = your_model, ...)` additionally exercises
  `er_simulate()`'s `sim_resp`.
- `er_plot_add_summary(model = your_model)` (paired with whichever
  summary builder matches what your `er_summary()` populates) exercises
  `er_summary()`.

If you only implement `er_predict()`, that's a complete, valid
implementation -- the other two are additive, not phases of a single
required rollout. And if a method genuinely can't do better than the
default (`NULL`), leaving it unimplemented is the correct choice, not a
gap to apologise for: every erplots builder that depends on
`er_simulate()`/`er_summary()` already has to handle "not available"
gracefully, since the default method returns `NULL` for every model
class that hasn't implemented one.

If your method requires an argument beyond the fixed contract (like
`toy_landmark`'s `landmark_time` above), remember that exercising it
means passing `predict_args`/`summary_args`/`simulate_args` explicitly
-- calling `er_plot_add_model(your_model)` with no `predict_args` will
correctly error if your method has no default for that argument, the
same way `toy_landmark`'s did above.

## Real-world implementations

The two examples above are toy classes built for this article. Two real
packages implement this interface for production modelling code, and are
worth reading directly if you want to see the pattern applied to a less
simplified model:

- [erglm](https://github.com/djnavarro/erglm) wraps `glm()`-based
  exposure-response models. Its `er_predict.erglm_model()` and
  `er_simulate.erglm_model()` methods follow the same shape as
  `toy_model`'s above (link-scale prediction with an inverse-link
  transform; a multivariate-normal draw over the coefficient vector for
  simulation) but handle all four families erglm supports
  (`binomial`/`poisson`/`gaussian`/`Gamma`), plus the auto-seed-and-report
  behaviour mentioned above. Its `er_summary.erglm_model()` populates all
  three keys at once (`p_value` for the exposure coefficient,
  `coefficients` for every term, `glance` for goodness-of-fit).
- [emaxnls](https://github.com/djnavarro/emaxnls) fits genuine
  Emax/sigmoidal models via `nls()`, and is the real-world analogue of
  `toy_emax` above: `er_summary.emaxnls()` always returns
  `p_value = NULL` and populates `coefficients` (one row per `E0`/`Emax`/
  `logEC50`) instead, for exactly the reason discussed above -- no single
  parameter is "the" effect. Its `emax_logistic()` variant fits a
  binary response but returns an object with class
  `c("emaxlogistic", "emaxnls")`, so it reuses the same three methods via
  ordinary S3 inheritance; the methods branch internally (on
  `inherits(model, "emaxlogistic")`) to keep predictions bounded in
  `[0, 1]` and to draw Bernoulli rather than Gaussian noise for
  `sim_resp`, rather than needing a fourth, separate set of methods.

Both are `Suggests`-only dependencies of erplots (see `DESCRIPTION`),
used here only as worked examples -- there's no requirement to depend on
either to implement this interface for your own model class.

## See also

- `?er_model_interface` for the terse, canonical statement of the
  contract this article expands on.
- [The plotting grammar](design.html) for how `er_predict()`/
  `er_simulate()`/`er_summary()`'s outputs flow into the layers that
  consume them (`config$predictions`, `config$summary`, `config$p_value`,
  and so on).
- [Extending erplots](extending.html) if, once your model works with the
  built-in layers, you also want to change *how* a layer draws that
  model's output -- a different kind of extensibility from the one this
  article covers.
