---
title: "Visual predictive checks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(erplots)
library(emaxnls)
library(erglm)
```

The mini-grammar for VPC plots is considerably simpler than the one for exposure-response plots.

## Binary response 

### By exposure

```{r}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

The default approach for binary outcomes is to bin the exposure variable by quartiles, and then show the confidence interval for the response rate for both observed and simulated: 

```{r}
erglm_data |> 
  er_vpc(exposure = aucss, response = ae1) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
``` 

### By continuous covariate

Continuous `plot_by` does not have to be the same as the exposure variable:

```{r}
erglm_data |> 
  er_vpc(exposure = aucss, response = ae1, plot_by = weight) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
``` 


### By discrete covariate

```{r}
mod <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_vpc(exposure = aucss, response = ae1, plot_by = sex) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
``` 

## Continuous response 

### By exposure

```{r}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

For continuous outcomes, more options are available. It is possible to adopt the same approach as we did for binary outcomes, and it does work:

```{r}
emax_df |> 
  er_vpc(exposure = exp_1, response = rsp_1) |> 
  er_vpc_add_observed() |> 
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
```

However, this approach only considers whether the model is correctly predicting the mean response within each bin. 

As an alternative, we can switch to a distributional approach:

```{r}
emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous"
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()
```

For more detailed examination, it is also possible to display nonparametric confidence intervals for the observed data quantiles, plotted (like the default mean/errorbar pair) at each bin's numeric median on the exposure scale.

```{r}
emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous"
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_errorbar
  ) |>
  plot()
```

### By continuous covariate

Continuous `plot_by` variables can be something other than the exposure:

```{r}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)

emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous",
    plot_by = cnt_a
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_errorbar
  ) |>
  plot()


emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous",
    plot_by = cnt_a
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()

```

### By discrete covariate

```{r}
mod <- erglm_model(
  biomarker_change ~ aucss + sex, erglm_data, family = gaussian()
)

erglm_data |> 
  er_vpc(exposure = aucss, response = biomarker_change, plot_by = sex) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()

```

## Troubleshooting plot legibility

A persistent problem when creating VPC plots is cleaning up an illegible plot. To an extent, this problem is unavoidable: VPC plots attempt to display a *lot* of information in a single image, and it has a tendency to make them difficult to read. Because the specific pattern of illegibility tends to vary from plot to plot, it is almost impossible to build in an automated fix to this problem. Instead of attempting an automated fix that will very likely not work, the `er_vpc()` mini-grammar exposes some customization tools that you can use to clean up a VPC plot that doesn't look very nice. In this section, we outline some of the options that you have. 

```{r}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

### Errorbar collisions

We'll start by creating a deliberately terrible VPC plot, using the Emax regression in the `mod` object. Suppose that, for one reason or another, you have been asked to show *five* quantiles for this model rather than the usual three. This is rarely a good idea because it leads to terrible crowding in the plot. Here's what happens if we do this, using errorbar styling for both the observed data and the model simulations: 

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_errorbar
  ) |>
  plot()
```

There's no simple fix for this: in this plot we have unpleasant collisions between the observed and simulated layers, *and* between the five individual `probs` within each layer. However, you do have some options. Assuming for the sake of this vignette that you really do need all five quantiles, and for the moment assuming that both the simulated and observed layers must be displayed in an errorbar style, there are three arguments we can tinker with to improve legibility:

- `dodge` shifts a whole builder's error bars sideways -- pair an equal and opposite `dodge` on the observed and simulated builders to pull the two layers apart.
- `prob_dodge_width` spreads a single builder's own `probs` apart, symmetrically around the bin's true position.
- `errorbar_width` narrows or widens the error bars themselves, giving the eye more room to separate bars that remain close together after dodging

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(
    style = er_style_vpc_observed_quantile_errorbar,
    dodge = -0.005, 
    prob_dodge_width = 0.005,
    errorbar_width = 0.0025
  ) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_errorbar,
    dodge = 0.005, 
    prob_dodge_width = 0.005,
    errorbar_width = 0.0025
  ) |>
  plot()
```

It takes a little bit of trial and error to find values that work in any specific case (and it does help to take a close look at the documentation to see how the different arguments are interpreted), but nevertheless this is a substantial improvement. It is not perfect, but it works reasonably well.

### Ribbon collisions

A different kind of legibility issue arises when ribbons are used to display confidence intervals. As before, we'll illustrate this using the extreme case when five quantiles are being displayed. A common visual style used for VPC plots is to use connected lines for the observed quantiles, and then ribbons to show model predictions and uncertainty at each quantile. In this case, however, the confidence bands are so closely packed together that they blend together visually and it is impossible to tell where the boundaries are:

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()
```

Sometimes you can improve legibility in this case by placing more emphasis on the edges of the bands, and reducing the salience of the band shading. The code below illustrates this, and it helps somewhat, but the plot is still difficult to read:

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_ribbon,
    ribbon_alpha = 0.1,
    ribbon_edges = TRUE
  ) |>
  plot()
```

Ultimately, in this situation the most likely resolution is that you would have to revert from five quantile bands to the usual three:

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.5, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_ribbon,
    ribbon_alpha = 0.2,
    ribbon_edges = TRUE
  ) |>
  plot()
```

Some of the distributional information is lost, but overall the plot is a lot easier to understand.