---
title: "Visual predictive checks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(erplots)
library(emaxnls)
library(erglm)
```

Alongside the `er_plot()` mini-grammar covered in the [binary](plot-binary.html), [continuous](plot-continuous.html), and [count](plot-count.html) articles, erplots supplies a second, smaller mini-grammar built around `er_vpc()` for constructing visual predictive checks (VPCs). A VPC compares what a model predicts against what was actually observed, binned by exposure (or some other variable of interest), so that any systematic mismatch between model and data is easy to spot. The grammar is deliberately narrower than `er_plot()`'s in one respect: there's no colour/facet precedence rule to reconcile across builders, since an optional `stratify_by` splits the plot into facet panels only (see `?er_vpc`), never colour. But the same model-agnostic philosophy still applies -- any model implementing `er_predict()`/`er_simulate()` (see [Implementing the model interface](model-interface.html)) can be visualised this way, and this article uses the same erglm and emaxnls models as the other articles to demonstrate it. As with `er_plot()`, `er_vpc()` itself only sets up the plot's variables and bins; nothing is drawn until `er_vpc_add_observed()` and `er_vpc_add_simulated()` are added and the pipeline is plotted.

## Binary response 

For a binary response, `er_vpc_add_observed()` bins the data and computes the observed response rate (plus a confidence interval) in each bin, and `er_vpc_add_simulated()` draws `nsim` replicate datasets from the model and summarises those the same way -- so the two rates can be compared directly, bin by bin.

### VPC by exposure

```{r}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

The default approach for binary outcomes is to bin the exposure variable by quartiles, and then show the confidence interval for the response rate for both observed and simulated: 

```{r}
erglm_data |> 
  er_vpc(exposure = aucss, response = ae1) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
``` 

### VPC by continuous covariate

The variable used to bin the data and position the x-axis, `plot_by`, defaults to the exposure variable, but it doesn't have to be. Setting it to a different continuous variable lets you check whether the model's predictions track the observed data as *that* variable changes, even when it isn't part of the fitted model itself. Here we bin by `weight` instead of `aucss`, using the same exposure-only model as above:

```{r}
erglm_data |> 
  er_vpc(exposure = aucss, response = ae1, plot_by = weight) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
``` 


### VPC by discrete covariate

The same idea applies to a categorical `plot_by`: there's no x-axis binning to do, since the categories already partition the data, and each level gets its own position. Here we bin by `sex`, using a model that actually includes `sex` as a covariate, so the simulated layer's predictions can differ between the two groups:

```{r}
mod <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_vpc(exposure = aucss, response = ae1, plot_by = sex) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
``` 

## Stratified panels

`plot_by` changes *what's binned* along the x-axis; `stratify_by` instead keeps the x-axis as-is and splits the plot into one facet panel per level of some other variable, via `ggplot2::facet_wrap()`. It's useful when you want to check the model against a covariate without giving up the usual exposure-response view. Unlike `plot_by`, `stratify_by` never changes what's plotted on the x-axis or how it's binned -- both `er_vpc_add_observed()` and `er_vpc_add_simulated()` behave exactly as before within each panel.

### Discrete `stratify_by`

A categorical `stratify_by` is used as-is, one panel per level. Here we reuse the `ae1 ~ aucss + sex` model from above, faceting by `sex` while keeping `aucss` on the x-axis:

```{r}
erglm_data |> 
  er_vpc(exposure = aucss, response = ae1, stratify_by = sex) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
```

### Continuous `stratify_by`

A numeric `stratify_by` is automatically split into `n_strata` quantile bins (4 by default), and `er_vpc()` reports a message when it does this, since it's a decision made on your behalf. Here we facet by `weight`, overriding the default with `n_strata = 3`:

```{r}
erglm_data |> 
  er_vpc(exposure = aucss, response = ae1, stratify_by = weight, n_strata = 3) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
```

`stratify_by` must resolve to a different variable than `plot_by` (which defaults to the exposure variable) -- faceting by the exact variable already driving the x-axis binning would give each panel a single bin, so `er_vpc()` errors rather than allow it.

## Continuous response 

A continuous response supports the same mean/errorbar comparison used above for binary outcomes, but it also supports a genuinely distributional comparison: checking whether the model reproduces the *shape* of the response distribution at each exposure level, not just its mean.

### VPC by exposure

```{r}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

We'll start by showing the limitations of the default "mean plus confidence interval" visual style. This is set as the default style for VPCs because it has the virtue of working regardless of whether the `plot_by` variable is continuous or discrete, and regardless of whether the `response` variable is binary, continuous or counts. But as you can see from the plot below, it's not the best choice for a continuous response variable:

```{r}
emax_df |> 
  er_vpc(exposure = exp_1, response = rsp_1) |> 
  er_vpc_add_observed() |> 
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
```

The plot renders correctly, but the only thing it shows is whether the model can correctly predict the mean response within each bin. Because of this, it's almost never the best choice when the `response` is continuous or count data.

As an alternative, we can switch to a genuinely distributional comparison: connected lines for several observed quantiles, and shaded ribbons showing the corresponding simulated quantiles. This style is the most commonly used approach when VPCs are applied for pharmacokinetic models, and works well for the exposure-response case with a continuous response variable:

```{r}
emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous"
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()
```

This works well, and is usually the best choice, but one thing that is missing in this plot is an expression of uncertainty about the observed quantiles. Sometimes that is useful to have, in which case you can do something like this, where the `ci_quantile()` function is used under the hood to construct confidence intervals for observed quantiles:

```{r}
emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous"
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_errorbar
  ) |>
  plot()
```

### VPC by continuous covariate

As with the binary case, a continuous `plot_by` doesn't have to be the exposure variable. Here the Emax model includes a covariate effect of `cnt_a` on the baseline parameter `E0`, so we bin on `cnt_a` directly to check the model's percentile predictions against it. Here is the VPC plotted in an errorbar style:

```{r}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)

emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous",
    plot_by = cnt_a
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_errorbar
  ) |>
  plot()
```

For comparison, the plot below shows the same VPC in the ribbon style. The two show the same underlying comparison styled differently; which one to use is mostly a matter of taste (and, as covered below, legibility):

```{r}
emax_df |> 
  er_vpc(
    exposure = exp_1, 
    response = rsp_1, 
    response_type = "continuous",
    plot_by = cnt_a
  ) |> 
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |> 
  er_vpc_add_simulated(
    model = mod, 
    seed = 1234, 
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()
```

### VPC by discrete covariate

As before, a categorical `plot_by` works with the default mean/errorbar pair too, this time using a model that includes `sex` as a covariate on the continuous response:

```{r}
mod <- erglm_model(
  biomarker_change ~ aucss + sex, erglm_data, family = gaussian()
)

erglm_data |> 
  er_vpc(exposure = aucss, response = biomarker_change, plot_by = sex) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  plot()
```

## Theming

`er_vpc_theme()` adjusts labels, titles, axis limits, and the overall ggplot2 theme, without changing which variable drives which aesthetic. Every argument defaults to `NULL` ("leave unchanged"), so repeated calls only touch the fields they supply:

```{r}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())

erglm_data |> 
  er_vpc(exposure = aucss, response = ae1) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod, seed = 1234) |>
  er_vpc_theme(
    xlab = "AUC at steady state",
    ylab = "Probability of event",
    title = "Visual predictive check",
    theme_base = ggplot2::theme_minimal()
  ) |>
  plot()
```

Note that `xlab` labels `plot_by` (the VPC's actual x-axis variable), not necessarily `exposure` -- the two only coincide when `plot_by` wasn't overridden. `er_vpc_theme()` doesn't cover everything: there's no argument for the observed/simulated colour scale, for instance, since it's fixed to keep the two aligned across builders that mix colour and fill for the same distinction. For anything not covered, `+ ggplot2::theme(...)`/`+ ggplot2::labs(...)` on the object returned by `plot()` remains the general-purpose escape hatch.

## Troubleshooting plot legibility

A persistent problem when creating VPC plots is cleaning up an illegible plot. To an extent, this problem is unavoidable: VPC plots attempt to display a *lot* of information in a single image, and it has a tendency to make them difficult to read. Because the specific pattern of illegibility tends to vary from plot to plot, it is almost impossible to build in an automated fix to this problem. Instead of attempting an automated fix that will very likely not work, the `er_vpc()` mini-grammar exposes some customization tools that you can use to clean up a VPC plot that doesn't look very nice. In this section, we outline some of the options that you have. 

```{r}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

### Errorbar collisions

We'll start by creating a deliberately terrible VPC plot, using the Emax regression in the `mod` object. Suppose that, for one reason or another, you have been asked to show *five* quantiles for this model rather than the usual three. This is rarely a good idea because it leads to terrible crowding in the plot. Here's what happens if we do this, using errorbar styling for both the observed data and the model simulations: 

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_errorbar) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_errorbar
  ) |>
  plot()
```

There's no simple fix for this: in this plot we have unpleasant collisions between the observed and simulated layers, *and* between the five individual `probs` within each layer. However, you do have some options. Assuming for the sake of this vignette that you really do need all five quantiles, and for the moment assuming that both the simulated and observed layers must be displayed in an errorbar style, there are three arguments we can tinker with to improve legibility:

- `dodge` shifts a whole builder's error bars sideways -- pair an equal and opposite `dodge` on the observed and simulated builders to pull the two layers apart.
- `prob_dodge_width` spreads a single builder's own `probs` apart, symmetrically around the bin's true position.
- `errorbar_width` narrows or widens the error bars themselves, giving the eye more room to separate bars that remain close together after dodging

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(
    style = er_style_vpc_observed_quantile_errorbar,
    dodge = -0.005, 
    prob_dodge_width = 0.005,
    errorbar_width = 0.0025
  ) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_errorbar,
    dodge = 0.005, 
    prob_dodge_width = 0.005,
    errorbar_width = 0.0025
  ) |>
  plot()
```

It takes a little bit of trial and error to find values that work in any specific case (and it does help to take a close look at the documentation to see how the different arguments are interpreted), but nevertheless this is a substantial improvement. It is not perfect, but it works reasonably well.

### Ribbon collisions

A different kind of legibility issue arises when ribbons are used to display confidence intervals. As before, we'll illustrate this using the extreme case when five quantiles are being displayed. A common visual style used for VPC plots is to use connected lines for the observed quantiles, and then ribbons to show model predictions and uncertainty at each quantile. In this case, however, the confidence bands are so closely packed together that they blend together visually and it is impossible to tell where the boundaries are:

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()
```

Sometimes you can improve legibility in this case by placing more emphasis on the edges of the bands, and reducing the salience of the band shading. The code below illustrates this, and it helps somewhat, but the plot is still difficult to read:

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.3, 0.5, 0.7, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_ribbon,
    ribbon_alpha = 0.1,
    ribbon_edges = TRUE
  ) |>
  plot()
```

Ultimately, in this situation the most likely resolution is that you would have to revert from five quantile bands to the usual three:

```{r}
emax_df |>
  er_vpc(
    exposure = exp_1,
    response = rsp_1,
    response_type = "continuous",
    n_bins = 5,
    probs = c(0.1, 0.5, 0.9)
  ) |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed = 1234,
    style = er_style_vpc_simulated_quantile_ribbon,
    ribbon_alpha = 0.2,
    ribbon_edges = TRUE
  ) |>
  plot()
```

Some of the distributional information is lost, but overall the plot is a lot easier to understand.