---
title: "Plotting: binary responses"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

erplots draws exposure-response plots from *any* model that implements
[er_model_interface]. This article uses a logistic regression model
fitted with erglm, but the plotting code itself has no knowledge of
`glm()`. It's the most detailed of the three response-type articles
(binary/continuous/count); the continuous and count articles link back
here for the model and group layers, which work identically
regardless of response type.

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

Unlike a plotting function that fits a model behind the scenes, erplots
expects you to fit the model yourself and pass it in explicitly:

```{r fit}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

## Defining plots

Basic usage

```{r basic-plot-1}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  plot()
```

Adding extra layers

```{r basic-plot-2}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_add_groups(group_by = aucss) |> 
  plot()
```

## Stratification

Stratification adds colour across all layers. This requires a model
that includes the stratification variable as a term:

```{r stratification-1}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |>
  plot()
```

You can suppress stratification for specific layers

```{r stratification-2}
erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  # keep_strata = FALSE needs a model that doesn't include the
  # stratification variable, so we pass the un-stratified `mod` here
  er_plot_add_model(mod, keep_strata = FALSE) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |>
  plot()
```

## Model layer

The default builder is `er_style_model_ribbonline()`, but you can also draw
spaghetti plots to represent parameter uncertainty with
`er_style_model_spaghetti()`. Spaghetti plots require the model to
implement `er_simulate()` (erglm's models do); models that only implement
`er_predict()` fall back to `er_style_model_ribbonline()` with a message.
This layer doesn't look at `response_type` at all -- it only consumes
[er_predict()]/[er_simulate()] output -- so everything in this section
applies unchanged to continuous and count responses too.

```{r model-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod, style = er_style_model_spaghetti) |> 
  er_plot_add_quantiles() |> 
  plot()
```

## Summary layer

`er_plot_add_summary()` annotates the base panel with a statistic
derived from the model (or, for `er_style_summary_n()`, purely
descriptive of the raw data) -- see [er_model_interface] for what a
model's own [er_summary()] method can return. The default builder,
`er_style_summary_pvalue()`, draws whatever headline p-value the model
reports, in whichever corner is furthest from the observed data:

```{r summary-pvalue}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_summary(model = mod) |> 
  plot()
```

`er_style_summary_gof()` draws a different kind of annotation --
goodness-of-fit rather than a p-value -- built from a curated, compact
subset of `er_summary()`'s `glance` field (`N`, `AIC`, `BIC`, `R²`,
whichever are present and non-`NA`). erglm's own models don't populate
`glance` yet, so this example defines a small demonstration method on
the spot purely to show the display idiom -- a real model package would
do this once, in its own `er_summary.<class>()` method, the same way
erglm already does for `p_value`:

```{r summary-gof}
er_summary.erglm_model_glance_demo <- function(model, ...) {
  list(glance = tibble::tibble(
    n = stats::nobs(model),
    aic = stats::AIC(model),
    bic = stats::BIC(model)
  ))
}
registerS3method("er_summary", "erglm_model_glance_demo", er_summary.erglm_model_glance_demo)

mod_glance_demo <- mod
class(mod_glance_demo) <- c("erglm_model_glance_demo", class(mod_glance_demo))

erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_summary(model = mod_glance_demo, style = er_style_summary_gof) |> 
  plot()
```

`er_style_summary_coefficients()` (one line per model parameter, for
models with no single privileged p-value) and `er_style_summary_n()` (a
model-agnostic observation count) are two further builders for this
layer -- see `?er_style_summary` for all four.

## Quantile layer

You can modify the number of bins:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles(bins = 6) |> 
  plot()
```

You can also modify the confidence level for the interval around each
bin's summary. For a binary response this is a Clopper-Pearson interval
for the response *rate*; see the
[continuous](plot-continuous.html#quantile-layer) and
[count](plot-count.html#quantile-layer) articles for how this layer
adapts its summary statistic and interval method to those response
types.

```{r quantile-2}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles(bins = 6, conf_level = .8) |> 
  plot()
```

## Data layer

`er_plot_add_data()` adds the raw observations. By default
(`er_style_data_overlay()`), points are drawn at their true `(exposure,
response)` coordinates in the *main* model panel -- for a binary response
this is a scatter with a small vertical jitter, since the y-values are
exactly 0/1 and would otherwise overplot into two solid lines:

```{r data-overlay-binary}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |> 
  plot()
```

### `er_style_data_overlay()` vs. `er_style_data_boxjitter()`

`er_style_data_boxjitter()` is the older, panel-based design, and is
binary-response only: it splits responders/non-responders into separate
panels above/below the main plot, each showing a boxplot of the
exposure values with the raw jittered points layered on top -- so the
panel shows the exposure *distribution* conditional on response, not
just individual points. There is no built-in panel-based builder for a
continuous/count response; `er_style_data_overlay()` (raw points at their
true `(exposure, response)` coordinates, shown in the
[continuous](plot-continuous.html#data-layer) and
[count](plot-count.html#data-layer) articles) covers that case
there, and a custom `"panel"`-layout builder (e.g. a single
color-encoded panel) remains possible via [er_style_tag()] if a project
needs one -- see `vignettes/articles/design.Rmd`'s "Extending erplots"
section. Each builder declares which of the two structural families it
belongs to via [er_style_tag()], which is what `er_plot_add_data()` uses
to decide whether to merge it into the main panel or stack it in panels
below.

Building both side by side (via patchwork's `|` operator) makes the
difference concrete:

```{r data-compare-binary, fig.width = 10}
p_overlay <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |> 
  er_plot_build()

p_boxjitter <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data(style = er_style_data_boxjitter) |> 
  er_plot_build()

p_overlay$output | p_boxjitter$output
```

Stratification looks the same for both: color/fill always means strata
for `er_style_data_boxjitter()`, sharing the model curve's own legend, the
same way `er_style_data_overlay()`'s color aesthetic does for any response
type:

```{r data-compare-stratified, fig.width = 10}
p_overlay_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_data() |> 
  er_plot_build()

p_boxjitter_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_data(style = er_style_data_boxjitter) |> 
  er_plot_build()

p_overlay_strat$output | p_boxjitter_strat$output
```

## Group layer

Multiple grouping variables are allowed. Like the model layer, this
layer doesn't look at `response_type` at all -- it only consumes the
exposure variable -- so everything in this section applies unchanged to
continuous and count responses too.

```{r group-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = c(aucss, sex)) |> 
  plot()
```

Stratification propagates to the group layer:

```{r group-2}
erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = aucss) |> 
  plot()
```

The default builder is `er_style_group_boxplot()`, but you can also use
violin plots with `er_style_group_violin()`:

```{r group-3}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = sex, style = er_style_group_violin) |> 
  plot()
```

## VPC plot

`er_vpc_plot()` is a model-agnostic VPC-style plot that compares observed
data against model-simulated data, operating on plain data frames rather
than an `er_plot` object. For a binary response it compares observed vs.
simulated response *rates*:

```{r vpc-1}
sim_binary <- erglm_vpc_sim(mod, seed = 5218)
er_vpc_plot(erglm_data, sim_binary, aucss, ae1, group_by = aucss)
```

See the [continuous](plot-continuous.html#vpc-plot) and
[count](plot-count.html#vpc-plot) articles for how this generalises to
means (with a t-interval or exact Poisson interval, respectively) for
those response types.
