---
title: "Plotting"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

erplots draws exposure-response plots from *any* model that implements the
[er_model_interface]. This article uses logistic regression models fitted
with erglm, but the plotting code itself has no knowledge of `glm()`.

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

Unlike a plotting function that fits a model behind the scenes, erplots
expects you to fit the model yourself and pass it in explicitly:

```{r fit}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

## Defining plots

Basic usage

```{r basic-plot-1}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  plot()
```

Adding extra components

```{r basic-plot-2}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_data() |>
  er_plot_show_groups(group_by = aucss) |> 
  plot()
```

## Stratification

Stratification adds colour across all components. This requires a model
that includes the stratification variable as a term:

```{r stratification-1}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |>
  plot()
```

You can suppress stratification for specific components

```{r stratification-2}
erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  # keep_strata = FALSE needs a model that doesn't include the
  # stratification variable, so we pass the un-stratified `mod` here
  er_plot_show_model(mod, keep_strata = FALSE) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |>
  plot()
```

## Model component

The default builder is `build_model_ribbonline()`, but you can also draw
spaghetti plots to represent parameter uncertainty with
`build_model_spaghetti()`. Spaghetti plots require the model to
implement `er_simulate()` (erglm's models do); models that only implement
`er_predict()` fall back to `build_model_ribbonline()` with a message.

```{r model-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod, builder = build_model_spaghetti) |> 
  er_plot_show_quantiles() |> 
  plot()
```

## Quantile component

You can modify the number of bins:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles(bins = 6) |> 
  plot()
```

You can also modify the confidence level for the interval around each
bin's summary. For a binary response this is a Clopper-Pearson interval
for the response *rate*:

```{r quantile-2}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles(bins = 6, conf_level = .8) |> 
  plot()
```

## Continuous responses

`er_plot()` auto-detects whether the response is binary (logical, or
values entirely in `{0, 1}`) or continuous, and you can override the
detection with `response_type`. The model component and quantile
component both adapt: quantile bins report a mean with a t-interval
instead of a rate with a Clopper-Pearson interval. The data layer
component has no continuous-response variant (see below), so it's
omitted here.

```{r continuous-1}
mod_gaussian <- erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())

erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_show_model(mod_gaussian) |> 
  er_plot_show_quantiles() |> 
  plot()
```

`er_vpc_plot()` generalises the same way, comparing observed vs.
simulated means rather than rates:

```{r continuous-2}
sim_gaussian <- erglm_vpc_sim(mod_gaussian, seed = 3947)
er_vpc_plot(erglm_data, sim_gaussian, aucss, biomarker_change, group_by = aucss)
```

Count (Poisson-style) responses, such as an adverse-event count, are
auto-detected as `"continuous"` and summarised the same way (bin mean
plus t-interval) unless you declare `response_type = "count"` explicitly:

```{r continuous-3}
mod_poisson <- erglm_model(ae_count ~ aucss, erglm_data, family = poisson())

erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_show_model(mod_poisson) |> 
  er_plot_show_quantiles() |> 
  plot()
```

Declaring `response_type = "count"` swaps the t-interval approximation
for an exact Poisson interval (bin mean plus [poisson_interval()]
instead of [t_interval()]), which never produces a negative lower bound
-- useful for low-count bins, where the t-interval approximation can. For
`erglm_data`'s own `ae_count`, none of the bin means are low enough for
this to actually happen, so the two plots above would look almost
identical if you re-ran the last one with `response_type = "count"`. To
make the difference concrete, here's a synthetic dataset where the
placebo arm has only 2 events among 20 subjects:

```{r count-synthetic}
set.seed(84)
placebo_counts <- rpois(20, 0.05)
while (sum(placebo_counts) != 2) placebo_counts <- rpois(20, 0.05)

aucss_dosed <- sort(runif(80, 100, 3000))
count_dosed <- rpois(80, 0.1 + 0.003 * aucss_dosed)

low_count_data <- data.frame(
  aucss = c(rep(0, 20), aucss_dosed),
  ae_count = c(placebo_counts, count_dosed)
)

mod_low_count <- erglm_model(ae_count ~ aucss, low_count_data, family = poisson())
```

The default (t-interval) path's placebo-arm error bar dips visibly below
zero -- a nonsensical negative event rate:

```{r count-1}
low_count_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_show_model(mod_low_count) |> 
  er_plot_show_quantiles() |> 
  plot() +
  ggplot2::coord_cartesian(ylim = c(-0.5, 1.5), xlim = c(-50, 800))
```

Declaring `response_type = "count"` keeps the same placebo-arm point
estimate but replaces the t-interval with an exact Poisson interval,
which stays non-negative:

```{r count-2}
low_count_data |> 
  er_plot(aucss, ae_count, response_type = "count") |> 
  er_plot_show_model(mod_low_count) |> 
  er_plot_show_quantiles() |> 
  plot() +
  ggplot2::coord_cartesian(ylim = c(-0.5, 1.5), xlim = c(-50, 800))
```

The same declaration works for `er_vpc_plot()`:

```{r count-3}
sim_poisson <- erglm_vpc_sim(mod_poisson, seed = 6142)
er_vpc_plot(
  erglm_data, sim_poisson, aucss, ae_count, group_by = aucss,
  response_type = "count"
)
```

## Data component

`er_plot_show_data()` adds the raw observations. By default
(`build_data_overlay()`), points are drawn at their true `(exposure,
response)` coordinates in the *main* model panel -- a plain scatter for a
continuous/count response, or a scatter with a small vertical jitter for
a binary response (whose y-values are exactly 0/1 and would otherwise
overplot into two solid lines). This works the same way for every
response type, with no special-casing needed:

```{r data-overlay-binary}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |> 
  plot()
```

```{r data-overlay-continuous}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_show_model(mod_gaussian) |> 
  er_plot_show_data() |> 
  plot()
```

### `build_data_overlay()` vs. `build_data_boxjitter()`

`build_data_boxjitter()` is the older, panel-based design, and is
binary-response only: it splits responders/non-responders into separate
panels above/below the main plot, each showing a boxplot of the
exposure values with the raw jittered points layered on top -- so the
panel shows the exposure *distribution* conditional on response, not
just individual points. There is no built-in panel-based builder for a
continuous/count response; `build_data_overlay()` (raw points at their
true `(exposure, response)` coordinates) covers that case, and a custom
`"panel"`-layout builder (e.g. a single color-encoded panel) remains
possible via [er_layout()] if a project needs one -- see
`vignettes/articles/design.Rmd`'s "Extending erplots" section. Each
builder declares which of the two structural families it belongs to via
[er_layout()], which is what `er_plot_show_data()` uses to decide
whether to merge it into the main panel or stack it in panels below.

Building both side by side (via patchwork's `|` operator) makes the
difference concrete for a binary response:

```{r data-compare-binary, fig.width = 10}
p_overlay <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |> 
  er_plot_build()

p_boxjitter <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data(builder = build_data_boxjitter) |> 
  er_plot_build()

p_overlay$output | p_boxjitter$output
```

Stratification looks the same for both: color/fill always means strata
for `build_data_boxjitter()` on a binary response, sharing the model
curve's own legend, the same way `build_data_overlay()`'s color
aesthetic does for any response type:

```{r data-compare-stratified, fig.width = 10}
p_overlay_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_data() |> 
  er_plot_build()

p_boxjitter_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_data(builder = build_data_boxjitter) |> 
  er_plot_build()

p_overlay_strat$output | p_boxjitter_strat$output
```

## Group component

Multiple grouping variables are allowed:

```{r group-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = c(aucss, sex)) |> 
  plot()
```

Stratification propagates to the group component:

```{r group-2}
erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = aucss) |> 
  plot()
```

The default builder is `build_group_boxplot()`, but you can also use
violin plots with `build_group_violin()`:

```{r group-3}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = sex, builder = build_group_violin) |> 
  plot()
```
