---
title: "Plotting"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

erplots draws exposure-response plots from *any* model that implements the
[er_model_interface]. This article uses logistic regression models fitted
with erglm, but the plotting code itself has no knowledge of `glm()`.

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

Unlike a plotting function that fits a model behind the scenes, erplots
expects you to fit the model yourself and pass it in explicitly:

```{r fit}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

## Defining plots

Basic usage

```{r basic-plot-1}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  plot()
```

Adding extra components

```{r basic-plot-2}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_data() |>
  er_plot_show_groups(group_by = aucss) |> 
  plot()
```

## Stratification

Stratification adds colour across all components. This requires a model
that includes the stratification variable as a term:

```{r stratification-1}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |>
  plot()
```

You can suppress stratification for specific components

```{r stratification-2}
erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  # keep_strata = FALSE needs a model that doesn't include the
  # stratification variable, so we pass the un-stratified `mod` here
  er_plot_show_model(mod, keep_strata = FALSE) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |>
  plot()
```

## Model component

Default is `style = "ribbonline"` but you can also draw spaghetti plots to
represent parameter uncertainty. Spaghetti plots require the model to
implement `er_simulate()` (erglm's models do); models that only implement
`er_predict()` fall back to `"ribbonline"` with a message.

```{r model-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod, style = "spaghetti") |> 
  er_plot_show_quantiles() |> 
  plot()
```

## Quantile component

You can modify the number of bins:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles(bins = 6) |> 
  plot()
```

You can also modify the confidence level for the interval around each
bin's summary. For a binary response this is a Clopper-Pearson interval
for the response *rate*:

```{r quantile-2}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles(bins = 6, conf_level = .8) |> 
  plot()
```

## Continuous responses

`er_plot()` auto-detects whether the response is binary (logical, or
values entirely in `{0, 1}`) or continuous, and you can override the
detection with `response_type`. The model component and quantile
component both adapt: quantile bins report a mean with a t-interval
instead of a rate with a Clopper-Pearson interval. The data layer
component has no continuous-response variant (see below), so it's
omitted here.

```{r continuous-1}
mod_gaussian <- erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())

erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_show_model(mod_gaussian) |> 
  er_plot_show_quantiles() |> 
  plot()
```

`er_vpc_plot()` generalises the same way, comparing observed vs.
simulated means rather than rates:

```{r continuous-2}
sim_gaussian <- erglm_vpc_sim(mod_gaussian, seed = 3947)
er_vpc_plot(erglm_data, sim_gaussian, aucss, biomarker_change, group_by = aucss)
```

Count (Poisson-style) responses, such as an adverse-event count, are
auto-detected as `"continuous"` and summarised the same way (bin mean
plus t-interval) unless you declare `response_type = "count"` explicitly:

```{r continuous-3}
mod_poisson <- erglm_model(ae_count ~ aucss, erglm_data, family = poisson())

erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_show_model(mod_poisson) |> 
  er_plot_show_quantiles() |> 
  plot()
```

Declaring `response_type = "count"` swaps the t-interval approximation
for an exact Poisson interval (bin mean plus [poisson_interval()]
instead of [t_interval()]), which never produces a negative lower bound
-- useful for low-count bins, where the t-interval approximation can. For
`erglm_data`'s own `ae_count`, none of the bin means are low enough for
this to actually happen, so the two plots above would look almost
identical if you re-ran the last one with `response_type = "count"`. To
make the difference concrete, here's a synthetic dataset where the
placebo arm has only 2 events among 20 subjects:

```{r count-synthetic}
set.seed(84)
placebo_counts <- rpois(20, 0.05)
while (sum(placebo_counts) != 2) placebo_counts <- rpois(20, 0.05)

aucss_dosed <- sort(runif(80, 100, 3000))
count_dosed <- rpois(80, 0.1 + 0.003 * aucss_dosed)

low_count_data <- data.frame(
  aucss = c(rep(0, 20), aucss_dosed),
  ae_count = c(placebo_counts, count_dosed)
)

mod_low_count <- erglm_model(ae_count ~ aucss, low_count_data, family = poisson())
```

The default (t-interval) path's placebo-arm error bar dips visibly below
zero -- a nonsensical negative event rate:

```{r count-1}
low_count_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_show_model(mod_low_count) |> 
  er_plot_show_quantiles() |> 
  plot() +
  ggplot2::coord_cartesian(ylim = c(-0.5, 1.5), xlim = c(-50, 800))
```

Declaring `response_type = "count"` keeps the same placebo-arm point
estimate but replaces the t-interval with an exact Poisson interval,
which stays non-negative:

```{r count-2}
low_count_data |> 
  er_plot(aucss, ae_count, response_type = "count") |> 
  er_plot_show_model(mod_low_count) |> 
  er_plot_show_quantiles() |> 
  plot() +
  ggplot2::coord_cartesian(ylim = c(-0.5, 1.5), xlim = c(-50, 800))
```

The same declaration works for `er_vpc_plot()`:

```{r count-3}
sim_poisson <- erglm_vpc_sim(mod_poisson, seed = 6142)
er_vpc_plot(
  erglm_data, sim_poisson, aucss, ae_count, group_by = aucss,
  response_type = "count"
)
```

## Data component

`er_plot_show_data()` adds the raw observations. By default (`style =
"overlay"`), points are drawn at their true `(exposure, response)`
coordinates in the *main* model panel -- a plain scatter for a
continuous/count response, or a scatter with a small vertical jitter for
a binary response (whose y-values are exactly 0/1 and would otherwise
overplot into two solid lines). This works the same way for every
response type, with no special-casing needed:

```{r data-overlay-binary}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data() |> 
  plot()
```

```{r data-overlay-continuous}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_show_model(mod_gaussian) |> 
  er_plot_show_data() |> 
  plot()
```

### `style = "overlay"` vs. `style = "jitter"`

`style = "jitter"` is the older, panel-based design: for a binary
response it splits responders/non-responders into separate jittered
panels above/below the main plot; for a continuous/count response it
draws a single panel (or one per stratum level) below the main plot,
with points colored continuously by the response value instead of split
into panels. Unlike overlay, its color aesthetic isn't available for
stratification on a continuous/count response, since the response value
already claims it.

Building both styles and placing them side by side (via patchwork's `|`
operator) makes the difference concrete for a binary response:

```{r data-compare-binary, fig.width = 10}
p_overlay <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data(style = "overlay") |> 
  er_plot_build()

p_jitter <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_data(style = "jitter") |> 
  er_plot_build()

p_overlay$output | p_jitter$output
```

and again for a continuous response, where `style = "jitter"` falls back
to a single response-colored panel below the model curve rather than an
overlay on it:

```{r data-compare-continuous, fig.width = 10}
p_overlay_cont <- erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_show_model(mod_gaussian) |> 
  er_plot_show_data(style = "overlay") |> 
  er_plot_build()

p_jitter_cont <- erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_show_model(mod_gaussian) |> 
  er_plot_show_data(style = "jitter") |> 
  er_plot_build()

p_overlay_cont$output | p_jitter_cont$output
```

Stratification also looks different between the two styles: overlay's
color aesthetic always means strata and shares the model curve's own
legend, regardless of response type, whereas jitter's color aesthetic
means strata only for a binary response (for continuous/count it falls
back to one titled panel per stratum level, each with its own response
colorbar):

```{r data-compare-stratified, fig.width = 10}
p_overlay_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_data(style = "overlay") |> 
  er_plot_build()

p_jitter_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_data(style = "jitter") |> 
  er_plot_build()

p_overlay_strat$output | p_jitter_strat$output
```

## Group component

Multiple grouping variables are allowed:

```{r group-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = c(aucss, sex)) |> 
  plot()
```

Stratification propagates to the group component:

```{r group-2}
erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = aucss) |> 
  plot()
```

The default is `style = "boxplot"` but you can also use violin plots:

```{r group-3}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = sex, style = "violin") |> 
  plot()
```
