---
title: "Plotting binary responses"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The erplots package supplies a mini-language for generating exposure-response plots commonly used in pharmacometric analyses. It is designed to be model agnostic, in the sense that it will work for any modelling tool that implements a few key interface functions (see [Implementing the model interface](model-interface.html)). It can support binary response data, continuous response data, and count response data. This article focuses on **binary data**, using a logistic regression model fitted using the erglm package. It covers the substantive aspects of plot construction, not the superficial features like labels, palettes, and visual theme. Those are covered by the [theming erplots](theming.html) article, which applies unchanged regardless of response type.

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

Unlike a plotting function that fits a model behind the scenes, erplots
expects you to fit the model yourself and pass it in explicitly:

```{r fit}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

## Defining plots

Basic usage

```{r basic-plot-1}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  plot()
```

Adding extra layers

```{r basic-plot-2}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_add_groups(group_by = aucss) |> 
  plot()
```

## Stratification

Stratification adds colour across all layers. This requires a model
that includes the stratification variable as a term:

```{r stratification-1}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |>
  plot()
```

You can suppress stratification for specific layers

```{r stratification-2}
erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  # keep_strata = FALSE needs a model that doesn't include the
  # stratification variable, so we pass the un-stratified `mod` here
  er_plot_add_model(mod, keep_strata = FALSE) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |>
  plot()
```

## Model layer

The default builder is `er_style_model_ribbonline()`, but you can also draw
spaghetti plots to represent parameter uncertainty with
`er_style_model_spaghetti()`. Spaghetti plots require the model to
implement `er_simulate()` (erglm's models do); models that only implement
`er_predict()` fall back to `er_style_model_ribbonline()` with a message.
This layer doesn't look at `response_type` at all -- it only consumes
`er_predict()`/`er_simulate()` output -- so everything in this section
applies unchanged to continuous and count responses too.

```{r model-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod, style = er_style_model_spaghetti) |> 
  er_plot_add_quantiles() |> 
  plot()
```

## Summary layer

`er_plot_add_summary()` annotates the base panel with a statistic
derived from the model (or, for `er_style_summary_n()`, purely
descriptive of the raw data) -- see [Implementing the model
interface](model-interface.html) for what a model's own `er_summary()`
method can return. The default builder,
`er_style_summary_pvalue()`, draws whatever headline p-value the model
reports, in whichever corner is furthest from the observed data:

```{r summary-pvalue}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_summary(model = mod) |> 
  plot()
```

`er_style_summary_gof()` draws a different kind of annotation --
goodness-of-fit rather than a p-value -- built from a curated, compact
subset of `er_summary()`'s `glance` field (`N`, `AIC`, `BIC`, `R²`,
whichever are present and non-`NA`). erglm's own models don't populate
`glance` yet, so the code below fakes up a version of `er_summary()`
that does, purely to demonstrate the display idiom. Ordinarily this is
something a model-fitting package writes once, permanently, for its
own kind of model (the same way erglm already has a permanent
`er_summary()` that supplies `p_value`); the chunk below just does that
same step by hand, inline, for demonstration purposes only -- it's not
something you'd normally write yourself when just using erplots:

```{r summary-gof}
er_summary.erglm_model_glance_demo <- function(model, ...) {
  list(glance = tibble::tibble(
    n = stats::nobs(model),
    aic = stats::AIC(model),
    bic = stats::BIC(model)
  ))
}
registerS3method("er_summary", "erglm_model_glance_demo", er_summary.erglm_model_glance_demo)

mod_glance_demo <- mod
class(mod_glance_demo) <- c("erglm_model_glance_demo", class(mod_glance_demo))

erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_summary(model = mod_glance_demo, style = er_style_summary_gof) |> 
  plot()
```

`er_style_summary_coefficients()` (one line per model parameter, for
models with no single privileged p-value) and `er_style_summary_n()` (a
model-agnostic observation count) are two further builders for this
layer -- see `?er_style_summary` for all four.

## Quantile layer

You can modify the number of bins:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles(bins = 6) |> 
  plot()
```

You can also modify the confidence level for the interval around each
bin's summary. For a binary response this is a Clopper-Pearson interval
for the response *rate*; see the
[continuous](plot-continuous.html#quantile-layer) and
[count](plot-count.html#quantile-layer) articles for how this layer
adapts its summary statistic and interval method to those response
types.

```{r quantile-2}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles(bins = 6, conf_level = .8) |> 
  plot()
```

## Data layer

`er_plot_add_data()` adds the raw observations. By default
(`er_style_data_overlay()`), points are drawn at their true `(exposure,
response)` coordinates in the *main* model panel -- for a binary response
this is a scatter with a small vertical jitter, since the y-values are
exactly 0/1 and would otherwise overplot into two solid lines:

```{r data-overlay-binary}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |> 
  plot()
```

### `er_style_data_overlay()` vs. `er_style_data_boxjitter()`

`er_style_data_boxjitter()` is a panel-based design, and is
binary-response only: it splits responders/non-responders into separate
panels above/below the main plot, each showing a boxplot of the
exposure values with the raw jittered points layered on top -- so the
panel shows the exposure *distribution* conditional on response, not
just individual points. There is no built-in panel-based builder for a
continuous/count response; `er_style_data_overlay()` (raw points at their
true `(exposure, response)` coordinates, shown in the
[continuous](plot-continuous.html#data-layer) and
[count](plot-count.html#data-layer) articles) covers that case
there, and a custom `"panel"`-layout builder (e.g. a single
color-encoded panel) remains possible via `er_style_tag()` if a project
needs one -- see the [Extending erplots](extending.html) article. Each
builder declares which of the two structural families it belongs to
via `er_style_tag()`, which is what `er_plot_add_data()` uses to decide
whether to merge it into the main panel or stack it in panels below.

Building both side by side (via patchwork's `|` operator) makes the
difference concrete:

```{r data-compare-binary, fig.width = 10}
p_overlay <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |> 
  er_plot_build()

p_boxjitter <- erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data(style = er_style_data_boxjitter) |> 
  er_plot_build()

p_overlay$output | p_boxjitter$output
```

Stratification looks the same for both: color/fill always means strata
for `er_style_data_boxjitter()`, sharing the model curve's own legend, the
same way `er_style_data_overlay()`'s color aesthetic does for any response
type:

```{r data-compare-stratified, fig.width = 10}
p_overlay_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_data() |> 
  er_plot_build()

p_boxjitter_strat <- erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_data(style = er_style_data_boxjitter) |> 
  er_plot_build()

p_overlay_strat$output | p_boxjitter_strat$output
```

## Group layer

Multiple grouping variables are allowed. Like the model layer, this
layer doesn't look at `response_type` at all -- it only consumes the
exposure variable -- so everything in this section applies unchanged to
continuous and count responses too.

```{r group-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = c(aucss, sex)) |> 
  plot()
```

Stratification propagates to the group layer:

```{r group-2}
erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_add_model(mod_strat) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = aucss) |> 
  plot()
```

The default builder is `er_style_group_boxplot()`, but you can also use
violin plots with `er_style_group_violin()`:

```{r group-3}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_add_model(mod) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = sex, style = er_style_group_violin) |> 
  plot()
```
