---
title: "Plotting: continuous responses"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

erplots draws exposure-response plots from *any* model that implements
[er_model_interface]. This article uses a gaussian model fitted with
erglm to cover continuous-response specifics; the model and group
components work identically for every response type, so this article
only shows their default usage and links to the
[binary responses](plot-binary.html) article for the builder-swapping
detail (spaghetti plots, violin plots).

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

```{r fit}
mod_gaussian <- erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())
```

`er_plot()` auto-detects whether a response is binary (logical, or
values entirely in `{0, 1}`) or continuous, and you can override the
detection with `response_type`. `biomarker_change` auto-detects as
`"continuous"`.

## Defining plots

```{r basic-plot-1}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_add_model(mod_gaussian) |> 
  er_plot_add_quantiles() |> 
  plot()
```

## Stratification

Stratification adds colour across all components, and requires a model
that includes the stratification variable as a term. See the
[binary responses](plot-binary.html#stratification) article for a fuller
worked example, including how to suppress stratification for specific
components with `keep_strata = FALSE`.

```{r stratification-1}
mod_gaussian_sex <- erglm_model(
  biomarker_change ~ aucss + sex, erglm_data, family = gaussian()
)

erglm_data |> 
  er_plot(aucss, biomarker_change, stratify_by = sex) |> 
  er_plot_add_model(mod_gaussian_sex) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |>
  plot()
```

## Model component

The model layer doesn't look at `response_type` at all -- it only
consumes [er_predict()]/[er_simulate()] output -- so it works exactly the
same way as for a binary response. See the
[binary responses](plot-binary.html#model-component) article for
`er_builder_model_spaghetti()` and the parameter-uncertainty rationale behind
it; the default builder is used here:

```{r model-1}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_add_model(mod_gaussian) |> 
  er_plot_add_quantiles() |> 
  plot()
```

## Quantile component

For a continuous response, each exposure-quantile bin is summarised by
its **mean** with a **t-interval**, rather than a rate with a
Clopper-Pearson interval:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_add_model(mod_gaussian) |> 
  er_plot_add_quantiles(bins = 6, conf_level = .8) |> 
  plot()
```

The t-interval is a reasonable default for a genuinely continuous
response like this one. See the [count responses](plot-count.html) article
for a case -- low-count Poisson data -- where the t-interval approximation
can misbehave (a negative lower bound), and the `response_type = "count"`
declaration that fixes it.

## Data component

`er_plot_add_data()` adds the raw observations at their true `(exposure,
response)` coordinates via `er_builder_data_overlay()`, the default and only
built-in builder for a continuous response -- no jitter is needed, since
the response isn't confined to 0/1:

```{r data-overlay-continuous}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_add_model(mod_gaussian) |> 
  er_plot_add_data() |> 
  plot()
```

There's no built-in panel-based alternative for a continuous response --
`er_builder_data_boxjitter()` (the older, panel-based responders/non-responders
design covered in the [binary responses](plot-binary.html#er_builder_data_overlay-vs--er_builder_data_boxjitter)
article) is binary-only. If you need a panel-based builder here (e.g. a
single color-encoded panel), you can write a custom one and tag it with
`er_builder_tag(fn, layout = "panel")` -- see `design.Rmd`'s "Extending erplots" section.

## Group component

The group layer doesn't look at `response_type` at all -- it only
consumes the exposure variable -- so it works exactly the same way as for
a binary response. See the
[binary responses](plot-binary.html#group-component) article for
multiple grouping variables and `er_builder_group_violin()`; the default
builder and a single grouping variable are shown here:

```{r group-1}
erglm_data |> 
  er_plot(aucss, biomarker_change) |> 
  er_plot_add_model(mod_gaussian) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = aucss) |> 
  plot()
```

## VPC plot

`er_vpc_plot()` generalises the same way as the quantile layer, comparing
observed vs. simulated **means** rather than rates:

```{r vpc-1}
sim_gaussian <- erglm_vpc_sim(mod_gaussian, seed = 3947)
er_vpc_plot(erglm_data, sim_gaussian, aucss, biomarker_change, group_by = aucss)
```
