---
title: "Plotting count responses"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The erplots package supplies a mini-language for generating exposure-response plots commonly used in pharmacometric analyses. It is designed to be model agnostic, in the sense that it will work for any modelling tool that implements a few key interface functions (see `?er_model_interface`). It can support binary response data, continuous response data, and count response data. This article focuses on **count data**, using a Poisson model fitted using the
erglm package. 

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

```{r fit}
mod_poisson <- erglm_model(ae_count ~ aucss, erglm_data, family = poisson())
```

Count responses, such as an adverse-event count, auto-detect as
`"continuous"` under `er_plot()`'s `"auto"` logic (they're neither
logical nor confined to `{0, 1}`), and are summarised the same way a
genuinely continuous response is (bin mean plus t-interval) unless you
declare `response_type = "count"` explicitly.

## Defining plots

```{r basic-plot-1}
erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_poisson) |> 
  er_plot_add_quantiles() |> 
  plot()
```

## Stratification

Stratification adds colour across all layers, and requires a model
that includes the stratification variable as a term. See the
[binary responses](plot-binary.html#stratification) article for a fuller
worked example, including how to suppress stratification for specific
layers with `keep_strata = FALSE`.

```{r stratification-1}
mod_poisson_sex <- erglm_model(
  ae_count ~ aucss + sex, erglm_data, family = poisson()
)

erglm_data |> 
  er_plot(aucss, ae_count, stratify_by = sex) |> 
  er_plot_add_model(mod_poisson_sex) |> 
  er_plot_add_quantiles() |> 
  er_plot_add_data() |>
  plot()
```

## Model layer

The model layer doesn't look at `response_type` at all -- it only
consumes `er_predict()`/`er_simulate()` output -- so it works exactly the
same way as for a binary response. See the
[binary responses](plot-binary.html#model-layer) article for
`er_style_model_spaghetti()`; the default builder is used here:

```{r model-1}
erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_poisson) |> 
  er_plot_add_quantiles() |> 
  plot()
```

## Summary layer

The summary layer doesn't look at `response_type` at all either -- it
only consumes whatever the model's own `er_summary()` method returns --
so it works exactly the same way as for a binary response. See the
[binary responses](plot-binary.html#summary-layer) article for
`er_style_summary_gof()` and the full four-builder set; the default
builder is used here:

```{r summary-1}
erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_poisson) |> 
  er_plot_add_summary(model = mod_poisson) |> 
  plot()
```

## Quantile layer

Under auto-detection, a count response is summarised the same way a
continuous response is -- bin mean plus t-interval:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_poisson) |> 
  er_plot_add_quantiles() |> 
  plot()
```

Declaring `response_type = "count"` swaps the t-interval approximation
for an exact Poisson interval (bin mean plus `ci_poisson()`
instead of `ci_t()`), which never produces a negative lower bound
-- useful for low-count bins, where the t-interval approximation can. For
`erglm_data`'s own `ae_count`, none of the bin means are low enough for
this to actually happen, so the two plots above would look almost
identical if you re-ran the last one with `response_type = "count"`. To
make the difference concrete, here's a synthetic dataset where the
placebo arm has only 2 events among 20 subjects:

```{r count-synthetic}
set.seed(84)
placebo_counts <- rpois(20, 0.05)
while (sum(placebo_counts) != 2) placebo_counts <- rpois(20, 0.05)

aucss_dosed <- sort(runif(80, 100, 3000))
count_dosed <- rpois(80, 0.1 + 0.003 * aucss_dosed)

low_count_data <- data.frame(
  aucss = c(rep(0, 20), aucss_dosed),
  ae_count = c(placebo_counts, count_dosed)
)

mod_low_count <- erglm_model(ae_count ~ aucss, low_count_data, family = poisson())
```

The default (t-interval) path's placebo-arm error bar dips visibly below
zero -- a nonsensical negative event rate:

```{r count-1}
low_count_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_low_count) |> 
  er_plot_add_quantiles() |> 
  plot() +
  ggplot2::coord_cartesian(ylim = c(-0.5, 1.5), xlim = c(-50, 800))
```

Declaring `response_type = "count"` keeps the same placebo-arm point
estimate but replaces the t-interval with an exact Poisson interval,
which stays non-negative:

```{r count-2}
low_count_data |> 
  er_plot(aucss, ae_count, response_type = "count") |> 
  er_plot_add_model(mod_low_count) |> 
  er_plot_add_quantiles() |> 
  plot() +
  ggplot2::coord_cartesian(ylim = c(-0.5, 1.5), xlim = c(-50, 800))
```

## Data layer

`er_plot_add_data()` adds the raw observations at their true `(exposure,
response)` coordinates via `er_style_data_overlay()`, the default and only
built-in builder for a count response -- no jitter is needed, since the
response isn't confined to 0/1:

```{r data-overlay-count}
erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_poisson) |> 
  er_plot_add_data() |> 
  plot()
```

There's no built-in panel-based alternative for a count response --
`er_style_data_boxjitter()` (the older, panel-based responders/non-responders
design covered in the [binary responses](plot-binary.html#er_style_data_overlay-vs--er_style_data_boxjitter)
article) is binary-only. If you need a panel-based builder here, you can
write a custom one and tag it with `er_style_tag(fn, layout = "panel")` -- see
the [Extending erplots](extending.html) article.

## Group layer

The group layer doesn't look at `response_type` at all -- it only
consumes the exposure variable -- so it works exactly the same way as for
a binary response. See the
[binary responses](plot-binary.html#group-layer) article for
multiple grouping variables and `er_style_group_violin()`; the default
builder and a single grouping variable are shown here:

```{r group-1}
erglm_data |> 
  er_plot(aucss, ae_count) |> 
  er_plot_add_model(mod_poisson) |> 
  er_plot_add_quantiles() |>
  er_plot_add_groups(group_by = aucss) |> 
  plot()
```
