---
title: "Time-to-event plots"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(erplots)
library(survival)
library(ertte)
```

Alongside the `er_plot()`/`er_vpc()` mini-grammars covered in the other articles, erplots supplies a third, separate mini-grammar built around `er_tte()` for Kaplan-Meier/survival-over-time figures. It has a different coordinate system to the other two: time on the x-axis, survival probability on the y-axis, rather than exposure-vs-response. `er_tte()` itself computes the Kaplan-Meier estimate once (`survival::survfit()`), and every layer added afterwards reads from that shared fit rather than recomputing it. As with `er_plot()`/`er_vpc()`, nothing is drawn until at least one `er_tte_add_*()` layer has been added and the pipeline is plotted.

This article uses the `lung` dataset from the survival package throughout, treating `status == 2` as the event indicator (the dataset codes `1` = censored, `2` = dead).

## A single curve

The minimal pipeline is `er_tte()` followed by `er_tte_add_curve()`, which draws the Kaplan-Meier step curve with a confidence band:

```{r}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  plot()
```

`er_tte_add_censor()` adds a tick mark at every censoring time, layered on top of the curve:

```{r}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_censor() |>
  plot()
```

`er_tte_add_risktable()` adds a number-at-risk panel, stacked below the curve via patchwork, with a shared x-axis:

```{r}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_censor() |>
  er_tte_add_risktable() |>
  plot()
```

## Stratified curves

Passing `stratify_by` to `er_tte()` splits the Kaplan-Meier estimate into one curve per level (`survival::survfit(... ~ strata)`), and every layer added afterwards follows suit. `lung$sex` is coded numerically (1/2), so we convert it to a factor first -- otherwise a numeric `stratify_by` is automatically quantile-binned rather than used as-is (useful for a genuinely continuous covariate, not what we want here):

```{r}
lung_sex <- lung |> transform(sex = factor(sex, labels = c("Male", "Female")))

lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_censor() |>
  er_tte_add_risktable() |>
  plot()
```

A numeric `stratify_by` is split into `n_strata` quantile bins instead, with a message reporting that this happened:

```{r}
lung |>
  er_tte(time, status == 2, stratify_by = age, n_strata = 3) |>
  er_tte_add_curve() |>
  plot()
```

### Log-rank test annotation

`er_tte_add_pvalue()` adds a corner-placed annotation of the log-rank test comparing survival across `stratify_by`'s levels (`survival::survdiff()`). It requires a stratified object -- calling it on an unstratified one errors, since a log-rank test needs at least two groups to compare:

```{r}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_pvalue() |>
  plot()
```

## Overlaying a parametric model

`er_tte_add_model()` overlays a fitted parametric $S(t)$ curve (with an uncertainty band) on top of the Kaplan-Meier estimate, for any model implementing `er_predict_survival()` (see [Implementing the model interface](model-interface.html)). This article uses `ertte`, the package that implements the full model interface for time-to-event models, fitting an accelerated failure time model:

```{r}
mod <- ertte_aft(Surv(time, status == 2) ~ sex, lung_sex)

lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_model(mod) |>
  plot()
```

`model` may reference covariates beyond the strata variable; erplots fills any other covariate from the plot data with a reference value (first factor level or numeric mean), the same approach `er_plot_add_model()` uses. A Cox model works the same way, and doesn't require `stratify_by` to be set at all:

```{r}
mod_cox <- ertte_coxph(Surv(time, status == 2) ~ age, lung)

lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_model(mod_cox) |>
  plot()
```

Since `age` isn't a stratification variable here, `er_tte_add_model()` predicts `S(t)` at `age`'s mean value from the plot data -- a single overlay curve, rather than one per stratum.

## Theming

`er_tte_theme()` adjusts labels, titles, axis limits, and the overall ggplot2 theme, without changing which variable drives which aesthetic. As with `er_plot_theme()`/`er_vpc_theme()`, every argument defaults to `NULL` ("leave unchanged"), so repeated calls only touch the fields they supply:

```{r}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_pvalue() |>
  er_tte_theme(
    xlab = "Days",
    ylab = "Overall survival",
    strata_lab = "Sex",
    title = "Kaplan-Meier estimate of overall survival",
    theme_base = ggplot2::theme_minimal()
  ) |>
  plot()
```

One theming argument is structural rather than purely cosmetic: `xlim` overwrites the time axis's limits directly, and those limits are what `er_tte_add_model()`'s default `time_grid` and `er_tte_add_risktable()`'s default `times` are computed from *at add-layer time*. So `er_tte_theme(xlim = ...)` needs to be called before those layers to affect their defaults -- calling it afterwards only changes the curve panel's own displayed range:

```{r}
lung |>
  er_tte(time, status == 2) |>
  er_tte_theme(xlim = c(0, 600)) |>
  er_tte_add_curve() |>
  er_tte_add_risktable() |>
  plot()
```

For anything `er_tte_theme()` doesn't cover, `+ ggplot2::theme(...)`/`+ ggplot2::labs(...)` remains the general-purpose escape hatch on the ggplot2 object `plot()` returns -- though note that once `er_tte_add_risktable()` is in play, the result is a patchwork composition of two panels rather than a single ggplot2 object, so such additions only apply to the top-level object, not automatically to both panels.
