---
title: "Time-to-event plots"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(erplots)
library(survival)
library(ertte)
```

Alongside the `er_plot()`/`er_vpc()` mini-grammars covered in the other articles, erplots supplies a third, separate mini-grammar built around `er_tte()` for Kaplan-Meier/survival-over-time figures. It has a different coordinate system to the other two: time on the x-axis, survival probability on the y-axis, rather than exposure-vs-response. `er_tte()` itself computes the Kaplan-Meier estimate once (`survival::survfit()`), and every layer added afterwards reads from that shared fit rather than recomputing it. As with `er_plot()`/`er_vpc()`, nothing is drawn until at least one `er_tte_add_*()` layer has been added and the pipeline is plotted.

This article uses the `lung` dataset from the survival package throughout, treating `status == 2` as the event indicator (the dataset codes `1` = censored, `2` = dead).

## A single curve

The minimal pipeline is `er_tte()` followed by `er_tte_add_curve()`, which draws the Kaplan-Meier step curve with a confidence band:

```{r single-curve}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  plot()
```

`er_tte_add_censor()` adds a tick mark at every censoring time, layered on top of the curve:

```{r curve-censor}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_censor() |>
  plot()
```

`er_tte_add_risktable()` adds a number-at-risk panel, stacked below the curve via patchwork, with a shared x-axis:

```{r curve-censor-risktable}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_censor() |>
  er_tte_add_risktable() |>
  plot()
```

## Stratified curves

Passing `stratify_by` to `er_tte()` splits the Kaplan-Meier estimate into one curve per level (`survival::survfit(... ~ strata)`), and every layer added afterwards follows suit. `stratify_by` must name a discrete/categorical variable -- `lung$sex` is coded numerically (1/2), so we convert it to a factor first:

```{r stratified-curve}
lung_sex <- lung |> transform(sex = factor(sex, labels = c("Male", "Female")))

lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_censor() |>
  er_tte_add_risktable() |>
  plot()
```

A genuinely continuous covariate, like `age`, needs binning into groups first -- `cut_quantile()` gives full control over bin count, tie-breaking, and labels:

```{r stratified-continuous-covariate}
lung |>
  transform(age_grp = cut_quantile(age, n = 3)) |>
  er_tte(time, status == 2, stratify_by = age_grp) |>
  er_tte_add_curve() |>
  plot()
```

### Summary annotation

`er_tte_add_summary()` adds a corner-placed text/label annotation. Its default style, `er_style_tte_summary_logrank()`, compares survival across `stratify_by`'s levels (`survival::survdiff()`); on an unstratified object, or one with only 1 stratum level present in the data, it simply draws nothing (a log-rank test needs at least two groups to compare) rather than erroring:

```{r logrank-pvalue}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_summary() |>
  plot()
```

Other builders don't depend on `stratify_by` at all. `er_style_tte_summary_n()` draws subject/event counts (one line per stratum when stratified, a single overall line otherwise):

```{r summary-n}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_summary(style = er_style_tte_summary_n) |>
  plot()
```

`er_style_tte_summary_coefficients()`/`er_style_tte_summary_gof()` draw from a fitted model's own [er_summary()] result instead, passed via `er_tte_add_summary()`'s `model` argument -- independent of whatever model, if any, was passed to `er_tte_add_model()`. See [Implementing the model interface](model-interface.html) for what a model needs to implement to support this.

## Overlaying a parametric model

`er_tte_add_model()` overlays a fitted parametric $S(t)$ curve (with an uncertainty band) on top of the Kaplan-Meier estimate, for any model implementing `er_predict_survival()` (see [Implementing the model interface](model-interface.html)). This article uses `ertte`, the package that implements the full model interface for time-to-event models, fitting an accelerated failure time model:

```{r model-aft}
mod <- ertte_aft(Surv(time, status == 2) ~ sex, lung_sex)

lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_model(mod) |>
  plot()
```

`model` may reference covariates beyond the strata variable; erplots fills any other covariate from the plot data with a reference value (first factor level or numeric mean), the same approach `er_plot_add_model()` uses. A Cox model works the same way, and doesn't require `stratify_by` to be set at all:

```{r model-cox}
mod_cox <- ertte_coxph(Surv(time, status == 2) ~ age, lung)

lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_model(mod_cox) |>
  plot()
```

Since `age` isn't a stratification variable here, `er_tte_add_model()` predicts `S(t)` at `age`'s mean value from the plot data -- a single overlay curve, rather than one per stratum.

## Theming

`er_tte_theme()` adjusts labels, titles, axis limits, and the overall ggplot2 theme, without changing which variable drives which aesthetic. As with `er_plot_theme()`/`er_vpc_theme()`, every argument defaults to `NULL` ("leave unchanged"), so repeated calls only touch the fields they supply:

```{r theming}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_summary() |>
  er_tte_theme(
    xlab = "Days",
    ylab = "Overall survival",
    strata_lab = "Sex",
    title = "Kaplan-Meier estimate of overall survival",
    theme_base = ggplot2::theme_minimal()
  ) |>
  plot()
```

`subtitle`/`caption` add further plot-level text, and `ylim` overrides the survival-probability axis's displayed range -- purely cosmetic, since a survival probability is always modelled on `[0, 1]` regardless of what's shown:

```{r theming-labels}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_theme(
    subtitle = "Lung cancer cohort",
    caption = "Source: survival::lung",
    ylim = c(0, 1.05)
  ) |>
  plot()
```

One theming argument is structural rather than purely cosmetic, unlike `ylim` above: `xlim` overwrites the time axis's limits directly, and those limits are what `er_tte_add_model()`'s default `time_grid` and `er_tte_add_risktable()`'s default `times` are computed from *at add-layer time*. So `er_tte_theme(xlim = ...)` needs to be called before those layers to affect their defaults -- calling it afterwards only changes the curve panel's own displayed range:

```{r theming-xlim}
lung |>
  er_tte(time, status == 2) |>
  er_tte_theme(xlim = c(0, 600)) |>
  er_tte_add_curve() |>
  er_tte_add_risktable() |>
  plot()
```

`format_p` controls how the log-rank annotation's p-value is displayed:

```{r theming-format-p}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_add_summary() |>
  er_tte_theme(format_p = scales::label_pvalue(accuracy = 0.0001, add_p = TRUE)) |>
  plot()
```

`draw_key` swaps the legend glyph the curve/model layers use, e.g. a point instead of the default filled rectangle:

```{r theming-draw-key}
lung_sex |>
  er_tte(time, status == 2, stratify_by = sex) |>
  er_tte_add_curve() |>
  er_tte_theme(draw_key = ggplot2::draw_key_point) |>
  plot()
```

`height_curve`/`height_risktable` set the relative heights patchwork gives to the two stacked panels once `er_tte_add_risktable()` is in play -- supplying only one leaves the other unchanged:

```{r theming-heights}
lung |>
  er_tte(time, status == 2) |>
  er_tte_add_curve() |>
  er_tte_add_risktable() |>
  er_tte_theme(height_curve = 3, height_risktable = 2) |>
  plot()
```

`format_percent` is also accepted and stored, reserved for a future risk-table/survival-probability builder that formats a value as a percentage -- no built-in TTE style currently reads it, so setting it has no visible effect today.

For anything `er_tte_theme()` doesn't cover, `+ ggplot2::theme(...)`/`+ ggplot2::labs(...)` remains the general-purpose escape hatch on the ggplot2 object `plot()` returns -- though note that once `er_tte_add_risktable()` is in play, the result is a patchwork composition of two panels rather than a single ggplot2 object, so such additions only apply to the top-level object, not automatically to both panels.
