---
title: "Plotting"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

erplots draws exposure-response plots from *any* model that implements the
[er_model_interface]. This article uses logistic regression models fitted
with erglm, but the plotting code itself has no knowledge of `glm()`.

```{r setup}
library(erplots)
library(erglm)
```

## Fit the model first

Unlike a plotting function that fits a model behind the scenes, erplots
expects you to fit the model yourself and pass it in explicitly:

```{r fit}
mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
```

## Defining plots

Basic usage

```{r basic-plot-1}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |> 
  plot()
```

Adding extra components

```{r basic-plot-2}
erglm_data |> 
  er_plot(exposure = aucss, response = ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_datastrip() |>
  er_plot_show_groups(group_by = aucss) |> 
  plot()
```

## Stratification

Stratification adds colour across all components. This requires a model
that includes the stratification variable as a term:

```{r stratification-1}
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())

erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_datastrip() |>
  plot()
```

You can suppress stratification for specific components

```{r stratification-2}
erglm_data |> 
  er_plot(
    exposure = aucss, 
    response = ae1, 
    stratify_by = sex
  ) |> 
  # keep_strata = FALSE needs a model that doesn't include the
  # stratification variable, so we pass the un-stratified `mod` here
  er_plot_show_model(mod, keep_strata = FALSE) |> 
  er_plot_show_quantiles() |> 
  er_plot_show_datastrip() |>
  plot()
```

## Model component

Default is `style = "ribbonline"` but you can also draw spaghetti plots to
represent parameter uncertainty. Spaghetti plots require the model to
implement `er_simulate()` (erglm's models do); models that only implement
`er_predict()` fall back to `"ribbonline"` with a message.

```{r model-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod, style = "spaghetti") |> 
  er_plot_show_quantiles() |> 
  plot()
```

## Quantile component

You can modify the number of bins:

```{r quantile-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles(bins = 6) |> 
  plot()
```

You can also modify the confidence level for the Clopper-Pearson interval
(this empirical-summary layer currently assumes a binary response):

```{r quantile-2}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles(bins = 6, conf_level = .8) |> 
  plot()
```

## Strip component


## Group component

Multiple grouping variables are allowed:

```{r group-1}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = c(aucss, sex)) |> 
  plot()
```

Stratification propagates to the group component:

```{r group-2}
erglm_data |> 
  er_plot(aucss, ae1, stratify_by = sex) |> 
  er_plot_show_model(mod_strat) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = aucss) |> 
  plot()
```

The default is `style = "boxplot"` but you can also use violin plots:

```{r group-3}
erglm_data |> 
  er_plot(aucss, ae1) |> 
  er_plot_show_model(mod) |> 
  er_plot_show_quantiles() |>
  er_plot_show_groups(group_by = sex, style = "violin") |> 
  plot()
```
