---
title: "Theming erplots"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of this article is to discuss the `er_plot_theme()` function, used to control the surface appearance of plots. In contrast to the `er_plot_add_*()` functions and their associated builder functions, which can change substantive features of the plot, the role of `er_plot_theme()` is to control labels, axis limits, palettes, plot titles, and other thematic aspects to the plot. 

```{r setup}
library(erplots)
library(erglm)

mod <- erglm_model(ae1 ~ aucss, erglm_data, family = binomial())
mod_strat <- erglm_model(ae1 ~ aucss + sex, erglm_data, family = binomial())
mod_gaussian <- erglm_model(biomarker_change ~ aucss, erglm_data, family = gaussian())
```

Every example below reuses the same stratified plot:

```{r base-plot}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  plot()
```

`er_plot_theme()` slots into the same pipeline anywhere after `er_plot()`
-- its arguments are read lazily at build time, so it doesn't matter
whether it comes before or after the layer functions. Every argument
defaults to `NULL`, meaning "leave whatever was set before unchanged".

## Labels

`xlab`/`ylab` relabel the exposure/response axes; `strata_lab` relabels
the stratification legend (this errors if `stratify_by` wasn't set in
`er_plot()` -- there's no legend to label):

```{r labels}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_theme(
    xlab = "Steady-state AUC",
    ylab = "Adverse event (1)",
    strata_lab = "Sex"
  ) |>
  plot()
```

## Plot-level text

`title`/`subtitle`/`caption` add the usual plot-level annotations, via
`patchwork::plot_annotation()`:

```{r title}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_theme(
    title = "Adverse event vs. exposure",
    subtitle = "Stratified by sex",
    caption = "Source: erglm_data"
  ) |>
  plot()
```

## Axis limits

`xlim`/`ylim` override the exposure/response axis limits that `er_plot()`
otherwise computes automatically from the data:

```{r limits}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_theme(ylim = c(-0.1, 1.1)) |>
  plot()
```

## Visual theme

`theme_base` swaps out the overall ggplot2 theme (default
`ggplot2::theme_bw()`); `theme_extra` replaces the small set of
additional tweaks erplots applies on top (by default, a panel border
plus `legend.position = "bottom"`). Supplying `theme_extra` **replaces**
that default rather than adding to it, so re-include anything you want
to keep:

```{r theme}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_theme(
    theme_base = ggplot2::theme_minimal(),
    theme_extra = ggplot2::theme(legend.position = "right")
  ) |>
  plot()
```

## Discrete color/fill palette

`color_discrete`/`fill_discrete` take a discrete ggplot2 scale object
(e.g. from `ggplot2::scale_color_brewer()` or
`ggplot2::scale_color_viridis_d()`) and apply it wherever `colour`/`fill`
is genuinely mapped to the stratification variable:

```{r palette}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  er_plot_theme(
    color_discrete = ggplot2::scale_color_brewer(palette = "Dark2"),
    fill_discrete = ggplot2::scale_fill_brewer(palette = "Dark2")
  ) |>
  plot()
```

`color_discrete`/`fill_discrete` are left alone wherever `colour`/`fill`
means something other than strata -- e.g. `er_style_data_hex()`'s
density fill, below.

## Stratified quantile spacing

`dodge_width` adjusts the horizontal separation between strata within
each quantile bin. This is a theme-level setting, because it controls
the layout of stratification across the quantile layer rather than the
look of any single builder.

```{r dodge-width}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_quantiles(style = er_style_quantile_errorbar) |>
  er_plot_theme(dodge_width = 0.15) |>
  plot()
```

## Continuous color/fill palette

`color_continuous`/`fill_continuous` are the symmetric counterpart,
scoped to aesthetics mapped to something continuous *other* than the
stratification variable. `er_style_data_hex()`'s bin-density `fill` is
the one built-in example (a continuous/count response's response-colored
data layer is the other, but there's currently no built-in "panel"-layout
builder for it -- see `vignettes/articles/extending.Rmd` for writing a
custom one).

Left unstyled, `er_style_data_hex()` already supplies its own default --
a light-grey-to-navy gradient that fades toward the panel background as
a cell's count approaches zero, rather than ggplot2's own default
mid-intensity blue:

```{r continuous-default}
erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(mod_gaussian, style = er_style_model_line) |>
  er_plot_add_data(style = er_style_data_hex) |>
  plot()
```

`fill_continuous` overrides that default wherever it's set:

```{r continuous-palette}
erglm_data |>
  er_plot(aucss, biomarker_change) |>
  er_plot_add_model(mod_gaussian, style = er_style_model_line) |>
  er_plot_add_data(style = er_style_data_hex) |>
  er_plot_theme(fill_continuous = ggplot2::scale_fill_viridis_c()) |>
  plot()
```

As with `color_discrete`/`fill_discrete`, `color_continuous`/
`fill_continuous` only ever touch the aesthetic role they name --
supplying `fill_continuous` here has no effect on a *discrete* `fill`
mapping elsewhere in the same plot (e.g. a stratified model ribbon's
`fill = strata`), and vice versa for `fill_discrete`.

## Formatters

`format_p`/`format_percent`/`format_number` control how the summary and
quantile layers format their labels -- typically a `scales::label_*()`
call:

```{r formatters}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles() |>
  er_plot_add_summary(model = mod) |>
  er_plot_theme(
    format_p = scales::label_pvalue(accuracy = .0001),
    format_percent = scales::label_percent(accuracy = .1)
  ) |>
  plot()
```

## Legend key glyph

`draw_key` controls the glyph ggplot2 draws in the legend, e.g. a point
instead of the default filled rectangle:

```{r draw-key}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_data() |>
  er_plot_theme(draw_key = ggplot2::draw_key_point) |>
  plot()
```

## Panel heights

`height_base`/`height_data`/`height_group` set the relative heights
patchwork gives to the main panel, any data panels, and any group panels
-- supplying only one leaves the other two unchanged:

```{r heights, fig.height = 8}
erglm_data |>
  er_plot(aucss, ae1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles() |>
  er_plot_add_groups(sex) |>
  er_plot_theme(height_group = 5) |>
  plot()
```

## Calling `er_plot_theme()` more than once

Every argument defaults to `NULL`, so repeated calls accumulate rather
than replace: each call only touches the arguments it actually supplies,
the same merging behaviour as `ggplot2::theme()` itself. This makes it
natural to build up theming incrementally, e.g. once for labels and
again later for the palette:

```{r accumulate}
erglm_data |>
  er_plot(aucss, ae1, stratify_by = sex) |>
  er_plot_add_model(mod_strat) |>
  er_plot_add_data() |>
  er_plot_theme(xlab = "Steady-state AUC") |>
  er_plot_theme(color_discrete = ggplot2::scale_color_brewer(palette = "Dark2")) |>
  plot()
```

## Where to next

* [The plotting grammar](design.html) explains the layer-composition
  rules `er_plot_theme()` deliberately doesn't touch.
* [Extending erplots](extending.html) shows how to write a custom
  `style` builder, if changing *what's drawn* -- rather than *how it
  looks* -- is what you need.
